---
title: ''
output: 
    bookdown::word_document2: 
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)



# Display items

## Tables



## Figures
fig_DAG <- 1

# Read in files
```

**Title:** Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

<!-- **Title 1:** Sensitivity of tree growth to drought in a seasonally dry tropical forest using long-term dendrometer band measurements

**Title 2:** Contrasting effects of species and environmental predictors on tree growth sensitivity to different/two droughts in a seasonally dry tropical forest

**Title 3:** Diversity of species- and environment-mediated drought responses in a seasonally dry tropical forest

**Title 4:** Long-term dendrometer band measurements reveal diverse species- and environment-mediated drought responses in a seasonally dry tropical forest

**Title 5:** Deciduousness may exacerbate/worsen/amplify tree growth sensitivity during wet season droughts in seasonally dry tropical forests -->

**Authors**  
```{r eval = TRUE, echo=FALSE, warning=FALSE}
# library(flextable)
# set_flextable_defaults(font.family = "Arial", font.size = 9)
team <- read.csv("display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)

# team <- read.csv("doc/display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)

# ft <- flextable(team)
# ft <- line_spacing(ft, space = 1, part = "all")
# ft <- autofit(ft)
# ft <- width(ft, 1, width = 1.75)
# ft <- width(ft, 2, width = 1.75)
# ft <- width(ft, 3, width = 3)
# ft
```

Krishna Anujan^1,2^*  ORCiD: `r team$ORCID[1]`  
Sean McMahon^2,3^  ORCiD: `r team$ORCID[2]`  
Sarayudh Bunyavejchewin^4^  ORCiD: `r team$ORCID[3]`  
Stuart J. Davies^3^  ORCiD: `r team$ORCID[4]`  
Helene C. Muller-Landau^3^  ORCiD : `r team$ORCID[5]`  
Nantachai Pongpattananurak^5^  ORCiD: `r team$ORCID[6]`  
Kristina Anderson-Teixeira^1,3^  ORCiD: `r team$ORCID[7]`  

**Affiliations:**

^1^Smithsonian's National Zoo and Conservation Biology Institute, Front Royal, VA, USA  
^2^Smithsonian Environmental Research Centre, Edgewater, MD, USA  
^3^Smithsonian Tropical Research Institute, Panama  
^4^National Park, Wildlife and Plant Conservation Department, Bangkok, Thailand  
^5^Kasetsart University, Chatuchak, Bangkok, Thailand  

*=corresponding author

**Running title:**

**Article Type:** Letter
 
**Number of Words in Abstract:** 173  
**Number of Words in Main Text:** 4989  
**Number of References:** 46 
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0  

**Corresponding author address:** anujank@si.edu

# Keywords {-}

# Statement of Authorship {-}

# Data Accessibility Statement {-}

# Conflict of Interest {-}

\newpage

# Abstract (150 words) {-}
Droughts are increasingly impacting forests worldwide, threatening their critical carbon sink. 
Tree species adaptations to hot and dry conditions -- for example, dry season deciduousness in tropical seasonal forests -- might confer drought resistance, but such resistance may be variable across microenvironments and drought events. 
We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry forest in <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework--->to assess tree growth responses to two climatically distinct droughts. 
Species and individuals exhibited a wide range of drought responses within each drought and inconsistent responses across droughts. 
Deciduous species were more sensitive to early wet-season drought than evergreen species, and this difference was amplified in wetter microclimates. 
Exposed trees were more sensitive, but presumed greater water access through larger size or wetter topographic positions buffered growth declines. 
Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.

\newpage

# Introduction {-}

Globally, intense drought is becoming more frequent, affecting forest demography, diversity, and  carbon (C) cycling [@ipcc_climate_2023]. 
However, our ability to predict drought impacts on tree woody growth and resultant ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].  

<!--PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
<!-- Increasing drought risk is threatening demographic processes in global forests and consequently their consistent contributions to the land carbon sink [IPCC].   -->
<!-- - Every year, woody growth of trees in tropical forests removes roughly `r round (150*1/6,0)` Gt CO~2~ from the atmosphere [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016], contributing to a net biomass increase of ~7 Gt CO~2-eq~ [@harris_global_2021] that offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@friedlingstein_global_2022].copied from our grant proposal - something like this is more specific and perhaps more compelling -->
<!-- - Moreover, woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].  
- However, tree growth is vulnerable to drought, the frequency and intensity of which are increasing across the much of tropics [@tao_increasing_2022]. <!--and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].-->
<!-- 
- With changing climate patterns across the global tropics, emergence of novel climatic regimes like "very extreme droughts" in recent decades [@dahinden_future_2017; @luo_tropical_2022], there is an urgent need to understand the drivers of this variation, to be able to predict and respond to forest drought responses.   
 --> 

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019]. 
While forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @refs ], droughts with different characteristics may in fact have very different effects on tree growth [@refs].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. 
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those occurring at different times relative to regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@refs].
In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky? Hulshof?]. 

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
Finally, the microenvironments that individual trees experience influence growth responses to drought. <!--leading to varied atmospheric and hydrological stresses with distinct ecological consequences.-->
Across horizontal axes, microenvironments defined by topographic and edaphic variation shape tree drought responses. 
While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021] or acclimatized [@refs].
Indeed, there are observations of trees near streams undergoing greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments.
Aboveground, tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems [@refs] and thereby potential to access more reliable water sources [@refs].
However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium<!--, as described by Darcy's law -->[@mcdowell_darcy_2015; @fernández-de-uña_role_inpress], or by trait differences across vertical axis [@vinod_thermal_2023].
Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.

<!--Whether drought characteristics interfere positively or negatively with these evolutionary and physiological strategies and whether it leads to consistent or heterogenous sensitivities across drought events is as yet unknown.  -->


<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and microenvironments, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats. 
We test this by examining individual tree growth during two ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand. 
Specifically, we test predictions that
i) both species and individuals respond differently to different droughts. 
ii) Deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought. 
iii) Exposed trees are more sensitive to drought, but water availability, either through larger size or topographic location, can buffer growth declines.

<!-- - **habitat** : water availability is a key driver of tropical tree growth [@wagner_water_2012]. 
- **leaf habit** : drought resistance is higher in deciduous species because of leaf strategies that minimise water loss during months of high vulnerability. Deciduous and evergreen species have differential sensitivity to drought [@desouza_drought_2020].    
- **leaf habit x exposure** : species canopy strategies along with their leaf habit could exacerbate or counteract drought vulnerability [@rahman_speciesspecific_2019]. Under high exposure, deciduous species are potentially more drought resistant than evergreen species, while the pattern could be less clear under low exposure.  
- **leaf habit x habitat** : tropical evergreen and deciduous species have different habitat preferences [@kunert_leaf_2021] that could affect their drought resistance.   -->


```{r table, echo=FALSE, warning=FALSE, message=FALSE}
# figure 0-----------------------------

# make a figure with summaries

library(flextable)
set_flextable_defaults(font.family = "Arial", font.size = 9)

sumtab <- read.csv("display/summary_table.csv", stringsAsFactors = FALSE, check.names = FALSE)


ft <- flextable(sumtab)
ft <- set_header_labels(ft,
  na = ""
)
ft <- hline(ft, i = c(1, 4, 7), part = "body")
ft <- line_spacing(ft, space = 1, part = "all")
ft <- autofit(ft)
# ft <- height(ft, height = 0.3)
ft <- bold(ft, i = c(1, 4, 7), bold = TRUE, part = "body")
ft <- width(ft, 1, width = 3.5)
# ft <- width(ft, 2, width = 1.75)
# ft <- width(ft, 3, width = 3)
ft <- set_caption(ft, caption = "Summary table of hypotheses tested, effects of variables on drought sensitivity and corresponding evidence")
ft
```

# Materials and Methods {-}

***Sites and data*** 

Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
This is a 50-hectare plot in mixed deciduous tropical forest, with a mean annual temperature of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_. 
Long-term analysis of tree-ring data shows that tree growth at Huai Kha Khaeng has been influenced by temperature and rainfall [@vlam_temperature_2014].  

![***Figure 1. Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot. For four daily climate variables dry days, precipitation, maximum temperature and VPD, panel a) represents rolling means for 30 days around the focal day for that year. Grey line represents the mean of this rolling mean from 2008 to 2019 and ribbons represent standard deviation around this mean. Panel b) represents anomalies of this rolling mean from the values across the timeseries. Values, therefore represents the number of standard deviations from the mean on a given day. c) shows Standardised Precipitation Evapotranspiration Index (SPEI) values for each month in 2010 and 2015. Across all panels, light blue background represents the average wet season, from May to October. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)

In the HKK plot, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following standard protocol [@muller-landau_metal_2008]. 
Trees were selected were based on the relative abundance of species in the forest, with more individuals banded from the abundant species (ranging from 230 trees to 1 tree per species), and spanning the size and habitat distribution within the plot. 
On each tree, the dendrometer band was installed at 1.4 m from the ground, a height of 10 cm above the regular point of measure for the whole plot census (typically 1.3 m), except on irregular trunks or buttresses when they were shifted above. 
All dendrobands were censused two times a year - in the wet and dry season - typically in June and January. 
During the census, the size of the window opening along the band was measured using Mitutoyo digital callipers (least count = 0.01 mm) and the diameter of the tree at the dendrometer band was measured using standard DBH tape.  

***Calculating annualised growth increments***

Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year for each tree. 
We first removed measurements with likely data entry errors because of misplaced decimals, potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
We used the raw window size measurements combined with the diameter measurement at installation $(dia_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[condit].
We used a direct measurement of $dia_{1}$ at the time of installation if it was available.
If the tape measurement at the first window size measurement was not available for any band, we excluded window size measurements until the first available diameter measurement. 
We then calculated annualised diameter increments for each individual for each year by finding the difference in diameter from the previous year ($inc = dia_{t} - dia_{t-1}$) and adjusting for the number of days between measurements ($inc_{annual} = inc * \frac{tdif}{365}$).  

To ensure that the data did not smooth over multiple years, we excluded increment measures with missing values for $dia_{t-1}$ and then excluded large measurement outliers (> 3 standard deviations from the mean across all observations). 
Tape measurements themselves are not precise enough for calculating drought sensitivity; error for each tape measure is 1mm, error for growth is at least 2mm. 
In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
However, tape measurements can be used to check for consistency in trends or identify outliers. 
We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement with these annualised increments calculated from tape measurements. 
Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals across 30 species.

![***Figure 2: Growth increment timeseries and drought sensitivity*** across trees and species in the Huai Kha Khaeng ForestGEO plot. a) Median of annualised growth increments across species (colours) and the whole plot (black) from 2009 - 2022 from late wet/early dry season dendroband measurements. 2010 and 2015 were moderate and very strong ENSO events respectively. b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2.png)

***Drought years and characteristics***

We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - by examining two drought indices and corroborating with plot-level ecological responses and expert information. 
Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event [REF]. 
<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected. -->
We also used the SPEIbase dataset [@vicente-serrano_multiscalar_2010] to calculate the Standardised Precipitation-Evapotranspiration Index, a drought severity index, calculated at the monthly scale.
Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in 2010 and 2015 as drought events. 
These years also corresponded with expert knowledge of major droughts that affected the region as well as with plot-level growth declines in the timeseries. 
<!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
<!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
We also note that we chose not to include 2020 in our analysis despite low growth, because a) declines in growth can occur due to various reasons which could have distinct ecological responses in species and individuals b) 2020 did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.    

To characterise these two droughts, we used climate data from a weather station near the plot and publicly available climate data for the region. 
We used daily weather data collected from 2001 to 2019 at the Forest Fire Research Station [CHECK WHERE], located ~20 km from the ForestGEO plot. 
We first calculated daily values of Vapour Pressure Deficit (VPD) using the equation $$
VPD = 0.6108 * e^{17.27 * \frac{t_{max}}{t_{max}+237.3}}* (1-\frac{relative.humidity}{100})
$$
We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 15 day periods across the whole year for each year.
Finally, we calculated daily rolling anomalies for each variable for 2010 and 2015 from the long-term daily expectation as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$   

***Variables for analysis***

For the two drought years, 2010 and 2015, we calculated growth sensitivity for each tree to each drought as a growth anomaly from its mean annual increment from calculated from dendrometer band window size.

We first calculated the mean annual increment for each tree as the mean of the annualised increments across the whole timeseries.
We then calculated sensitivity for each tree $i$ for each drought as 
$$
sensitivity_{i} = \frac{drought.year.increment_{i} - mean.annual.increment_{i}}{mean.annual.increment_{i}}
$$
By scaling the difference in increment to the mean annual growth of each tree, this metric allows us to compare across slow and fast growing individuals and species. 

To account for potential biases in dendrometer band increments from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel_are_2016]. 
Timing of measurement and seasonal dynamics in tree hydration and bark dynamics could contribute to negative values in dendrometer band increments, unlike woody growth from tree ring datasets [@chitra-tarak_it_2015]. 
Using a zero-growth model, we calculated growth occurrence as 1 under positive diameter change and any change <=0 as 0. 
To account for biases introduced by timing of measurements and seasonal dynamics, we used this simplifying assumption to corroborate the patterns observed with the sensitivity metric.  

Across all individuals in the plot, we calculated topographic wetness index for an expectation of the water availability across space based on the concavity/convexity of a location. 
We used a Digital Elevation Model from a 5 m x 5 m elevation layer collected at plot establishment.
We used the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to fill and breach this layer, calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$.
We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
We estimated crown exposure for each tree at each census using crown illumination index, an ordinal categorical variable from 1 to 5 (least to most exposed) measured at each dendroband census.

We calculated deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
We defined deciduousness as the mean proportion of canopy loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous.  
Data was unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
An alternate measure of deciduousness is the duration spent at maximum crown loss.
This data was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for a larger list of species.  
<!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
We calculated niche breadth as the standard deviation of TWI across all observed individuals of each species in the whole plot census in 2008.  -->


***Statistical methods***

We first modelled sensitivity as a simple random intercept model

$$
sensitivity \sim 1 + (1 | Species)
$$ 

We then used the predicted sensitivities from this model and tested their correlations with deciduousness using simple linear models.

To model the interactive effects of deciduousness and wetness on sensitivity, we modelled

$$
sensitivity \sim 1 + TWI + deciduousness + TWI:decidousness
$$

We then calculated predictions across all combinations of values from observed maximum to minimum of TWI and deciduousness within this dataset.

To further disentange the microenvironmental drivers of drought sensitivity, we simultaneously modelled:

$$
sensitivity \sim DBH + CII + TWI + (DBH+CII+TWI | Species)
$$ 
and 
$$
CII \sim DBH
$$

Simultaneously modelling the effect of variables on sensitivity and the effect of DBH on CII in this way allows us to disentangle causual linkages between these processes. 
We chose to model a species random effect on all slopes here to account for variation across species instead of specific traits which may have different effects with each of the variables. 
However, we also ran simpler models with species random effects only on the intercept, the results of which are reported in the Supplementary Material. 
Across all models, we used Gaussian priors for the distribution of the response variable sensitivity, considered CII as an monotonic predictor (ordered factor) and centred and scaled the other predictors. 
We modelled CII as an ordinal categorical variable with a cumulative logit link function. 
In the combined form, the model therefore used a multivariate distribution of Gaussian and logit. 
<!-- Finally, to test full mediation effects, or the hypothesis that DBH effects on sensitivity are fully explained by its effects on exposure, we tested the same models, removing the DBH as a predictor of sensitivity. 
We then compared the fits of the partial mediation and full mediation models using Leave-One-Out Cross Validation.   -->

All statistical analyses were performed using R version 4.4.0 [@rcoreteam_language_2024]. 
Mixed models were fit through Bayesian regressions using MCMC methods using the package _brms_ [@bürkner_brms_2017].
We ran 4 MCMC chains for 3000 iterations with 1000 of these set as warmup.
We extracted draws, calculated distributions and checked convergence using functions from the _posterior_ package [@vehtari_ranknormalization_2021; @bürkner_posterior_2023].


# Results {-}

## Drought characteristics {-}

The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig 1).
In 2010, drought characteristics of lower precipitation, higher number of dry days, higher maximum temperature, higher VPD than the long-term mean and SPEI values below -1 occurred between February and April, in the dry season. 
On the other hand, in 2015, similar deviations occurred in May and June, in the wet season. 
In 2015, SPEI in May was -2, an anomalously low value that was not observed in any month in the entire dendroband census time period, suggesting a severe drought. 

```{r data summaries, echo=F, warning=F, message=F}
library(tidyverse)
tree.time <- readRDS("../data/HKK-dendro/sensitivity_data_formodels.RData")
# tree.time <- readRDS("data/HKK-dendro/sensitivity_data_formodels.RData")


tree.time.sp <- tree.time %>%
  group_by(spfull) %>%
  dplyr::summarise(
    inc_annual_med = round(median(inc_annual, na.rm = T), 2),
    inc_annual_sd = round(sd(inc_annual, na.rm = T), 2)
  ) %>%
  arrange(inc_annual_med)

sens.sp <- tree.time %>%
  filter(yr %in% c(2010, 2015)) %>%
  group_by(spfull, yr) %>%
  dplyr::summarise(
    sens_med = round(median(sens.prop, na.rm = T), 2),
    sens_mean = round(mean(sens.prop, na.rm = T), 2),
    sens_sd = round(sd(sens.prop, na.rm = T), 2)
  ) %>%
  arrange(sens_med)

# spread this for each year

sens.sp.wide <- sens.sp %>%
  ungroup() %>%
  dplyr::select(spfull, yr, sens_med) %>%
  pivot_wider(names_from = yr, values_from = sens_med, names_prefix = "sens.")

# paired correlation of 2010 and 2015 sensitivities across species

sens.sp.cor <- cor.test(sens.sp.wide$sens.2010, sens.sp.wide$sens.2015)
```

## Growth timeseries and Drought sensitivity {-}

The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig 2a). 
Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual, na.rm=T),2)` cm. 
Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed. 

Drought years were associated with lower growth across the plot. 
The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years. 
In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years. 
While there were more species differences in median responses to the 2010 drought, all species had lower increment than a typical year in 2015. 
Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4). 
Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`. The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%.

Trees on average had negative drought sensitivities for both droughts with considerable variation around it (Fig 2b). 
Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`. 
In 2010, *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` respectively. 
In 2015, *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` respectively. 


![***Figure 3: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010 and 2015. a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values. Line shown for significant correlation. b) Predicted sensitivity values across the observed range of deciduousness and topographic wetness index from a model with TWI, deciduosness and their interaction across all trees modelled for each drought year separately. For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png)

```{r model summaries, echo=F, warning=F, message=F}
# load data--------------------------
datasets <- readRDS("../data/HKK-dendro/sensitivity_data.RData")

tree.time <- datasets$tree.time
tree_vars <- datasets$tree_vars
sp_vars <- datasets$sp_vars

colnames(sp_vars)[1] <- "Species"

coefs_df <- readRDS("../results/models/non_negative/sensitivity_model_intercept.RData")
ranef_df <- coefs_df %>% filter(grepl("r_Species", param))
ranef_df$Species <- gsub("r_Species\\[|\\,Intercept\\]", "", ranef_df$param)

# add intercept to ranef_df
# make intercept vector
intercepts <- coefs_df %>%
  filter(param %in% "Intercept") %>%
  dplyr::select(median) %>%
  pull(median)
intercepts <- rep(intercepts, each = nrow(ranef_df) / 2)
ranef_df <- ranef_df %>%
  dplyr::mutate(
    intercept = intercepts + median,
    lwr = intercepts + lwr,
    upr = intercepts + upr
  )

# plot intercepts against species characteristics

# join ranef_df with sp_vars
ranef_df <- merge(ranef_df, sp_vars, by = "Species", all.x = TRUE)

# lms
coefs_dec_lms <- ranef_df %>%
  # filter(yr %in% yrs) %>%
  filter(Species != "ALPHVE") %>%
  nest_by(yr) %>%
  # dplyr::mutate(mod = list(lm(intercept ~ williams_dec, data = data))) %>%
  dplyr::mutate(mod = list(cor.test(data$williams_dec, data$intercept))) %>%
  dplyr::reframe(broom::tidy(mod))


isocline_coefs <- readRDS("../results/models/non_negative/coefs_isoclines_nore.RDS")
# isocline_coefs <- readRDS("results/models/non_negative/coefs_isoclines_nore.RDS")

isocline_coefs <- do.call(rbind, isocline_coefs)
isocline_coefs_dec <- isocline_coefs %>% filter(param == "b_williams_dec")
isocline_coefs_twi <- isocline_coefs %>% filter(param == "b_twi")
isocline_coefs_inter <- isocline_coefs %>% filter(param == "b_twi:williams_dec")
# isocline_coefs_twi
# head(isocline_coefs)

coefs_spre <- readRDS("../results/models/orderedcii/coefs_rel_spre.RDS")
# coefs_spre <- readRDS("results/models/orderedcii/coefs_rel_spre.RDS")
coefs_df_spre <- do.call(rbind, coefs_spre)
coefs_spre_twi <- coefs_df_spre %>% filter(param == "b_sensprop_twi_scaled")
# head(coefs_spre_twi)
```



## Species effects on drought sensitivity {-}

Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig 3a). 
Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S5), in alignment with median sensitivities across all trees. 
In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`). 
On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).  

## Microenvironment effects drought sensitivity {-}

The effect of wetness on sensitivity was different in 2010 and 2015, with 2015 effects interacting with deciduousness. 
Individual tree responses to drought were not correlated with each other (Fig S6). 
Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig 3b). 

The opposite effects of TWI on sensitivity in the wet and dry season drought, and the influence of deciduousness was also confirmed by models that controlled for crown exposure and size as well as species variation in intercept and slopes (Fig 4a). 
These models showed small negative sensitivities in wetter sites in the dry season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but strong positive effect of wetness in the wet season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`). 
The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig 4b). 
Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not show increase with wetness (Fig 4c). 
On the other hand, neither the mean sensitivities across species nor their response to TWI was associated with deciduousness in 2010.  

![***Figure 4: Drivers of variation in sensitivity among individuals***. All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes. a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree. b) wetness effect across all species and its relationship with deciduousness value. Line represents a significant correlation. c) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions. Black line represents overall relationship; coloured lines for each species reflects deciduousness values. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)

```{r exposure models, echo=F, warning=F, message=F}
coefs_spre_cii <- coefs_df_spre %>% filter(param == "bsp_sensprop_mocii_min1")
coefs_spre_ordcii <- coefs_df_spre %>% filter(param == "b_orderedciimin1_calcDBH_min1_scaled")
```

Crown exposure was associated with stronger negative sensitivities in both droughts.
In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 the negative effect was not as strong (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`). 
Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 (Fig 5b).

Accounting for effects of exposure, remaining effects of size on sensitivity differed between the two droughts. 
DBH was a significant predictor of exposure category, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a). 
Controlling for this direct effect of exposure, the remaining effect of size *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig 4a, Fig 5a). 


![***Figure 5: Effects of size, exposure and water availability*** from combined models. Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010 and b) 2015. Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size. c) Modelled sensitivity across crown exposure classes in 2010 and 2015. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png)

# Discussion {-}

<!---CONCLUSIONS---->
We show strong evidence that tropical tree drought sensitivity can be jointly and interactively shaped by drought characteristics, species adaptations and local environmental conditions (Table 1).
Analysing two droughts in a 14-year dendrometer band timeseries, we found that tropical tree species and individuals grew less than average in drought years, reflecting in lower plot-level growth (Fig 2). 
Comparing across droughts, we found that despite the *shorter duration of drought conditions*<!--careful here - didn't this continue into 2016?--> and greater absolute extremes of temperature, vapor pressure deficit, and lack of precipitation (Fig. 2a), the wet-season drought elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig 3). 
Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig 3b, Fig 4b). 
Further, besides consistent negative effects of crown exposure, we found that the drivers of variation among species and individuals in drought sensitivity were dissimilar between the drought events (Fig 4a). 
Water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts - larger trees (with likely deeper roots) in the dry-season drought and trees in wetter sites in the wet-season drought suffered smaller declines in growth (Fig 5). 
Taken together, our analysis of tree growth in two different drought years, and spanning meaningful variation, shows that species and individuals can respond differently to different drought events. 
Despite stand-level growth declines in drought years, we show that these broad patterns belie enormous heterogeneity in individual and species responses to drought in tropical forests with possible consequences for the maintenance of species diveristy and ecosystem stability.   

<!----Limitations in filling gap---->
Our analysis spans representative spatial and ecological variation, but we are limited in quantitatively analysing differences among droughts and variation in drought characteristics, because of low sample size. 
Given the diversity of factors that can influence tree growth and drought sensitivity in tropical tree communities, our analysis of 30 species across size classes and habitats spans this ecological and spatial (environmental) variation realistically and beyond the scope of previous such analyses. 
<!-- Long-timescale analyses with tree rings are limited by the fact that species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments. 
Tape measurements of growth from long-term ecological dynamics plots can be representative of species strategies but lack precision to parse signal from noise; error for each tape measure is 1mm, error for growth is at least 2mm while median annual growth across trees in HKK is ~1.4 mm and measuring deviation from this annual growth needs more precision.  -->
<!-- Traditional methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  -->
Further, although estimates of woody growth from dendroband timeseries can have measurement biases because of stem water shrinkage and bark peeling [@chitra-tarak_it_2015], our results are robust to simplifying assumptions that account for these biases (Fig S--).
However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts. 
Thus, quantitative analyses parsing out the specific effects of drought characteristics like relative influence of timing and degree of anomaly are currently impossible, limiting interpretations. 
In temperate forests, long-term records from tree ring analyses could disentangle species and individual sensitivities to differing drought characteristics, as annual rings are formed by a majority of the species.
Indeed @mcgregor_tree_2021 found that species sensitivities were different across drought events. 
However, as this accumulating dendrometer data record continues to grow, so does the potential to answer these questions across a wider range of forested ecosystems.  

<!-- ALT: We analysed species traits as static, but we might be missing intraspecific and plastic variation in drought sensitivity traits, as well as other axes of trait variation that might explain variation in drought sensitivity.
- Many hydraulic-related traits vary with tree height [ @vinod_thermal_2023], including the frequency of dry season deciduous leaf loss -- both within species and at the community level [@condit_ref; @meakem].  *caveat for species-level traits*  
- While leaf deciduousness was the main axis of hydraulic conservative strategy that we considered, leaf-level stomatal regulation and belowground root loss etc. are other potential strategies for conservative water use, and our interpretation of species patterns may be limited by our understanding of these processes across species.  
- Surprisingly, although deciduousness may have had an advantage in the dry season drought, because this strategy -> conservative water use in the dry season, species differences in this drought were not predicted by deciduousness.   -->

<!----Limitations in generalisation---->
<!-- ALT: Differential allocation to growth may not mean negative outcomes, it might mean demographic trade-offs
- Differences in allocation to growth may not be the complete picture - may have altered allocation to mortality or reproduction.  
- Some evidence that both understory and emergent species can have reduced survival during drought [@machado_forest_2023].  *for where else might they be allocating to?*
- 2015 ENSO drought changed regeneration dynamics in Thailand [@nutiprapun_effects_2023]
- also NSCs and then compensatory growth
- deciduous species have higher survival under multi-year droughts [@aguirre-guiterrez refs] -->

Tropical dry forests represent periodic hot/dry conditions, creating selective pressures for adaptative variation towards drought resistance, but this might not hold in other bioclimatic zones where adaptive pressures and thus ecological variability is different. 
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies and has resulted in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022) but it may not apply to the wet or everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   

<!----Contributions beyond---->
Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across drought events. 
Our finding that a wet season drought had bigger impact calls into question the generality of the findings of @zuidema_tropical_2022, which found that in species with tree-ring chronologies, dry season drought was most impactful. 
(Care must be taken in interpretation, however, as the 2015-2016 wet season ENSO drought was stronger (Fig 1)). 
<!-- Given the limitations of tree ring methods and tape measurements in measuring accurately measuring growth differences across a range of species in the tropics, methods so far could have led to capturing only the most dramatic droughts that affected (dry-associated) tree ring species and fast-growing individuals.  -->
*Afzelia xylocarpa*, one of the few annual-ring-forming species in HKK is noteworthy with the most distinct response across the two events; it had the highest (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought (Fig S5). 
More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. 
In contrast, our analyses reveal a range of responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies may result in heterogeneous consequences. 
Distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought in alignment with recent global analyses (@liu_drought_2022).    

<!----Science is better now---->
Drought sensitivity is often defined as a static species trait but our study reveals expanded dimensions of drought responses with possible consequences for management and restoration. 
Despite net decreased growth at the plot level (in agreement with landscape-scale in @aguirre-gutiérrez_functional_2022 and @bennett_sensitivity_2023), we show evidence that the same strategies are not uniformly successfully across spatiotemporal climatic variation. 
Static traits are often used to explain and predict forest responses to drought, especially in the tropics where long-term records are few [@guillemot_small_2022; @vico_how_2017a]. 
<!--aligns with remote sensing based inferences across global forests @liu_drought_2022-->
More inclusive frameworks for drought responses like @trugman_why_2021 (on mortality) encompass ecological complexity and covarying traits with environment like, but they assume consistent sensitivity of individual trees across events.
Our results, however, demonstrate an additional temporal dimension of variation, at least partially attributable to drought characteristics. 
Our results indicate that drought responses in tropical trees may not be a fixed response across events, but jointly shaped by species adaptations and local environmental conditions. 
Given these inconsistencies, diverse stands that may represent greater variation in ecological and evolutionary strategies, may have mechanisms for stable functioning under future droughts.  


<!---Other points--->

<!-- - (BCI light x size interaction in @rüger_growth_2011)   -->



# Acknowledgements {-}

We acknowledge the continued contributions of the Huai Kha Khaeng forest monitoring team in collecting data and maintaining the dendrometer bands. 
Ekaphan Kraichak and Kanisorn Chowtiwuttakorn helped with procuring elevation and climate datasets. 
José Medina-Vega, Valentine Herrmann and Eugenie Más provided input on improving analyses. 
KA was supported by the Smithsonian Institution Postdoctoral Fellowship and the ForestGEO Analytical Workshop, Asia 2024 in completing this work. 

\newpage
# References  {-}

