---
title: ''
output: 
    bookdown::word_document2: 
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document: 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
csl: new-phytologist.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)



# Display items

## Tables



## Figures

# Read in files

```


**Title:** Vulnerability of tree growth to precipitation in Thailand

**Authors**  
```{r eval = TRUE, echo=FALSE, warning=FALSE, eval=F}
library(flextable)
set_flextable_defaults(font.family = "Arial", font.size = 9)
team<- read.csv("display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)

ft <- flextable(team)
ft <- line_spacing(ft, space = 1, part = "all")
ft <- autofit(ft)
ft <- width(ft, 1, width = 1.75)
ft <- width(ft, 2, width = 1.75)
ft <- width(ft, 3, width = 3)
ft  

```


\newpage

# Abstract {-}


\newpage

# Introduction {-}

Tree woody growth key is a key component of aboveground productivity and affects the global carbon carbon cycle, but the drivers of its variation are poorly understood.
- Woody growth contributes x% of the aboveground woody productivity
- Long-lasting effects on the terrestrial carbon pool
- Intrinsic and extrinsic factors affecting variation.



# Methods {-}

ForestGEO data [@anderson-teixeira_ctfsforestgeo_2015]

***Sites and data*** - 


# Results {-}

# Discussion {-}




\newpage
# References  {-}

