---
title: ''
output: 
    bookdown::word_document2: 
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)



# Display items

## Tables



## Figures
fig_droughts <- 1
fig_drought_responses <- 2
fig_species <- 3
fig_microenvs <- 4
fig_dag <- 5

# Read in files
```

**Title:** Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

<!-- **Title 1:** Sensitivity of tree growth to drought in a seasonally dry tropical forest using long-term dendrometer band measurements

**Title 2:** Contrasting effects of species and environmental predictors on tree growth sensitivity to different/two droughts in a seasonally dry tropical forest

**Title 3:** Diversity of species- and environment-mediated drought responses in a seasonally dry tropical forest

**Title 4:** Long-term dendrometer band measurements reveal diverse species- and environment-mediated drought responses in a seasonally dry tropical forest

**Title 5:** Deciduousness may exacerbate/worsen/amplify tree growth sensitivity during wet season droughts in seasonally dry tropical forests -->

**Authors**  
```{r eval = TRUE, echo=FALSE, warning=FALSE}
# library(flextable)
# set_flextable_defaults(font.family = "Arial", font.size = 9)
team <- read.csv("display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)

# team <- read.csv("doc/display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)

# ft <- flextable(team)
# ft <- line_spacing(ft, space = 1, part = "all")
# ft <- autofit(ft)
# ft <- width(ft, 1, width = 1.75)
# ft <- width(ft, 2, width = 1.75)
# ft <- width(ft, 3, width = 3)
# ft
```

Krishna Anujan^1,2^*  ORCiD: `r team$ORCID[1]`  
Sean McMahon^2,3^  ORCiD: `r team$ORCID[2]`  
Sarayudh Bunyavejchewin^4^  ORCiD: `r team$ORCID[3]`  
Stuart J. Davies^3^  ORCiD: `r team$ORCID[4]`  
Helene C. Muller-Landau^3^  ORCiD : `r team$ORCID[5]`  
Nantachai Pongpattananurak^5^  ORCiD: `r team$ORCID[6]`  
Kristina Anderson-Teixeira^1,3^  ORCiD: `r team$ORCID[7]`  

**Affiliations:**

^1^Smithsonian's National Zoo and Conservation Biology Institute, Front Royal, VA, USA  
^2^Smithsonian Environmental Research Centre, Edgewater, MD, USA  
^3^Smithsonian Tropical Research Institute, Panama  
^4^National Park, Wildlife and Plant Conservation Department, Bangkok, Thailand  
^5^Kasetsart University, Chatuchak, Bangkok, Thailand  

*=corresponding author

**Running title:**

**Article Type:** Letter
 
**Number of Words in Abstract:** 150  
**Number of Words in Main Text:** 5200  
**Number of References:** 46  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0  

**Corresponding author address:** anujank@si.edu

# Keywords {-}

# Statement of Authorship {-}

# Data Accessibility Statement {-}

# Conflict of Interest {-}

\newpage

# Abstract (150 words) {-}
An observed increase in the frequency and severity of droughts threatens forest health worldwide, with important implications for this critical carbon sink. 
Tree species adaptations to hot and dry conditions--for example, dry season deciduousness in tropical seasonal forests--might confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events. 
We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry forest in Northern <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework---> to assess tree growth responses to two climatically distinct droughts. 
Species and individuals exhibited a wide range of growth responses within each drought, but these responses differed in intensity and affect  across droughts. 
Deciduous species were more sensitive to wet-season drought than evergreen species, and wetter microclimates amplified this difference. 
Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines. 
Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.

\newpage

# Introduction {-}

Globally, intense drought is becoming more frequent, affecting forest demography, diversity, and  carbon (C) cycling [@ipcc_climate_2023]. 
However, our ability to predict drought impacts on tree woody growth and the resultant ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].  

<!--PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
<!-- Increasing drought risk is threatening demographic processes in global forests and consequently their consistent contributions to the land carbon sink [IPCC].   -->
<!-- - Every year, woody growth of trees in tropical forests removes roughly `r round (150*1/6,0)` Gt CO~2~ from the atmosphere [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016], contributing to a net biomass increase of ~7 Gt CO~2-eq~ [@harris_global_2021] that offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@friedlingstein_global_2022].copied from our grant proposal - something like this is more specific and perhaps more compelling -->
<!-- - Moreover, woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].  
- However, tree growth is vulnerable to drought, the frequency and intensity of which are increasing across the much of tropics [@tao_increasing_2022]. <!--and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].-->
<!-- 
- With changing climate patterns across the global tropics, emergence of novel climatic regimes like "very extreme droughts" in recent decades [@dahinden_future_2017; @luo_tropical_2022], there is an urgent need to understand the drivers of this variation, to be able to predict and respond to forest drought responses.   
 --> 

<!-- From Sean: I like the detail of this proposed paragraph. It almost operates as a first paragraph? I guess the difference is going Ecology picture of droughts -> drought importance in to the carbon cycle to 'how to detect a drought vs. big picture droughts, ways it can affect ecology, to how you would measure ... :::: either way, there is some redundancy between these two paragraphs. -->

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019]. 
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may in fact have very different effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_ecology_letters_cryptic_phenology] [@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those occurring at different times relative to regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->

<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010]. 

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
Finally, the microenvironments that individual trees experience influence growth responses to drought. <!--leading to varied atmospheric and hydrological stresses with distinct ecological consequences.-->
Across horizontal axes, microenvironments defined by topographic and edaphic variation shape tree drought responses. 
While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted (e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021) or acclimatized [@refs].
Indeed, there are observations of trees near streams undergoing greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments.
Aboveground, tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and therefore the potential to access more reliable water sources [@stahl_depth_2013]. 
<!-- Sean::: I need to look at this source. Is this true? it doesn't really make sense unless coupled with other traits (e.g., deciduousness shifts accross ontogeny) as 'putting down a tap root' is one of the first things a seedling of a deep rooted species does (i.e., it's not allometry but strategy) -->

However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcys_2015; @fernández-de-uña_role_2023], or by trait differences across vertical axis [@vinod_thermal_2023].
Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.

<!--Whether drought characteristics interfere positively or negatively with these evolutionary and physiological strategies and whether it leads to consistent or heterogenous sensitivities across drought events is as yet unknown.  -->


<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and microenvironments, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats. 
Specifically, we test predictions that:
i) both species and individuals respond differently to different droughts;
ii) deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought; and,
iii) exposed trees are more sensitive to drought, but water availability, either through larger topographic location or tree size (and corresponding rooting volume), can buffer growth declines.
We tested these hypothses using  a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand. With data on 1820 trees across 30 species we built Bayesian regression models in a causal modelling framework to account for potential confounding relationships among the variables. This analysis adds important insight into the complexity of seasonal dry forest response to drought in  species-rich tropical forests. 

<!-- - **habitat** : water availability is a key driver of tropical tree growth [@wagner_water_2012]. 
- **leaf habit** : drought resistance is higher in deciduous species because of leaf strategies that minimise water loss during months of high vulnerability. Deciduous and evergreen species have differential sensitivity to drought [@desouza_drought_2020].    
- **leaf habit x exposure** : species canopy strategies along with their leaf habit could exacerbate or counteract drought vulnerability [@rahman_speciesspecific_2019]. Under high exposure, deciduous species are potentially more drought resistant than evergreen species, while the pattern could be less clear under low exposure.  
- **leaf habit x habitat** : tropical evergreen and deciduous species have different habitat preferences [@kunert_leaf_2021] that could affect their drought resistance.   -->


```{r table, echo=FALSE, warning=FALSE, message=FALSE}
# figure 0-----------------------------

# make a figure with summaries

library(flextable)
set_flextable_defaults(font.family = "Arial", font.size = 9)

sumtab <- read.csv("display/summary_table.csv", stringsAsFactors = FALSE, check.names = FALSE)


ft <- flextable(sumtab)
ft <- set_header_labels(ft,
  na = ""
)
ft <- hline(ft, i = c(1, 4, 7), part = "body")
ft <- line_spacing(ft, space = 1, part = "all")
ft <- autofit(ft)
# ft <- height(ft, height = 0.3)
ft <- bold(ft, i = c(1, 4, 7), bold = TRUE, part = "body")
ft <- width(ft, 1, width = 3.5)
# ft <- width(ft, 2, width = 1.75)
# ft <- width(ft, 3, width = 3)
ft <- set_caption(ft, caption = "Summary table of hypotheses tested, effects of variables on drought sensitivity and corresponding evidence. For each variable x year combination, \"+\" indicates a positive effect, \"-\" indicates a negative effect and \"n.s.\" indicates that the effect was not statistically significant.")
ft
```

# Materials and Methods {-}

***Sites and data*** 

Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
This is a 50-hectare plot in mixed deciduous tropical forest, with a mean annual temperature of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_. 
For analysis, we focussed on 30 of the most abundant species in the plot.   
Long-term analysis of tree-ring data shows that tree growth at HKK has been influenced by temperature and rainfall [@vlam_temperature_2014].  

![***Figure `r fig_droughts`. Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot. For four daily climate variables dry days, precipitation, maximum temperature and VPD, panel a) represents rolling means for 30 days around the focal day for that year. Grey line represents the mean of this rolling mean from 2008 to 2019 and ribbons represent standard deviation around this mean. Panel b) represents anomalies of this rolling mean from the values across the timeseries. Values, therefore represents the number of standard deviations from the mean on a given day. c) shows Standardised Precipitation Evapotranspiration Index (SPEI) values at four different time scales for each month in 2010 and 2015. A _k_-month SPEI integrates over _k_ months ending in the month current month. Across all panels, light blue background represents the average wet season, from May to October. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1_new.png)

In HKK, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following standard protocol [@muller-landau_metal_2008]. 
Trees were selected were based on the relative abundance of species in the forest, with more individuals banded from the abundant species (ranging from 230 trees to 1 tree per species), and spanning the size and habitat distribution within the plot. 
On each tree, the dendrometer band was installed at 1.4 m from the ground, a height of 10 cm above the regular point of measure for the whole plot census (typically 1.3 m), except on irregular trunks or buttresses when they were shifted above. 
All dendrobands were censused two times a year - in the wet and dry season - typically in June and January. 
During the census, the size of the window opening along the band was measured using Mitutoyo digital callipers (least count = 0.01 mm) and the diameter of the tree at the dendrometer band was measured using standard DBH tape.  

***Calculating annualised growth increments***

Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year for each tree. 
We first removed measurements with likely data entry errors because of misplaced decimals, potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
We used the raw window size measurements combined with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@condit_dendrometer_].
We used a direct measurement of $d_{1}$ at the time of installation if it was available.
If the tape measurement at the first window size measurement was not available for any band, we excluded window size measurements until the first available diameter measurement. 
We then calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($tdif$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{tdif}{365}$).  

To ensure that the data did not smooth over multiple years, we excluded increment measures with missing values for $d_{t-1}$ and then excluded large measurement outliers (> 3 standard deviations from the mean across all observations). 
Tape measurements themselves are not precise enough for calculating drought sensitivity; error for each tape measure is 1 mm, error for growth is at least 2 mm. 
In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
However, tape measurements can be used to check for consistency in trends or identify outliers. 
We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement with these annualised increments calculated from tape measurements. 
Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals across 30 species.

![***Figure `r fig_drought_responses`: Growth increment timeseries and drought responses*** across trees and species in the Huai Kha Khaeng ForestGEO plot. a) Median of annualised growth increments across ten dominant species (colours) and all trees (black) from 2009 - 2022. b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2.png)

***Drought years and characteristics***

We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses. 
Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event (Oceanic Niño Index @bamston_documentation_1997). 
<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected. -->
We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6- and 12- month scales using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig. `r fig_droughts`) and plot-level growth declines in the timeseries (Fig. `r fig_drought_responses`). 
<!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
<!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.    

To characterise the two focal droughts, we used data from a weather station near the plot and publicly available climate data for the region. 
We used daily weather data on maximum temperate, relative humidity and total precipitation collected from 2001 to 2019 at the Forest Fire Research Station <!--[CHECK WHERE]-->, located ~20 km from HKK, and calculated an additional variables vapour pressure deficit (VPD, @raesch_pvldcurve_2020). 
<!-- *(SPECIFY WHICH VARIABLES WERE OBTAINED FROM WHICH SOURCE, AND OTHER SOURCES.)* -->
<!-- We first calculated daily values of Vapour Pressure Deficit (VPD) using the equation $$ -->
<!-- VPD = 0.6108 * e^{17.27 * \frac{t_{max}}{t_{max}+237.3}}* (1-\frac{relative.humidity}{100}) -->
<!-- $$ -->
We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30 day periods across the whole timeseries.
Finally, we calculated daily rolling anomalies for each variable for 2010 and 2015 from the long-term daily expectation as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$   

***Variables for analysis***

For the two drought years, 2010 and 2015, we calculated growth sensitivity for each tree to each drought as a growth anomaly from its mean annual increment from calculated from dendrometer band window size.

We first  calculated the mean annual increment $\bar{a.inc_{i}}$ for each tree from annualised increment $ainc_{i,t}$<!-- as $\frac{\[ 
  \sum_{t}^{}{ainc_{i,t}} 
  \]}{\sum{t}}$-->.
We then calculated sensitivity for each tree $i$ for each drought year $y$ as 
$$
sensitivity_{i, t=y} = \frac{ainc_{i, t=y} - \bar{a.inc_{i}}}{\bar{a.inc_{i}}}
$$
By scaling the difference in increment to the mean annual growth of each tree, this metric allows us to compare across slow and fast growing individuals and species. 

<!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->
<!-- To account for potential biases in dendrometer band increments from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel_are_2016]. 
Timing of measurement and seasonal dynamics in tree hydration and bark dynamics could contribute to negative values in dendrometer band increments, unlike woody growth from tree ring datasets [@chitra-tarak_it_2015]. 
Using a zero-growth model, we calculated growth occurrence as 1 under positive diameter change and any change <=0 as 0. 
To account for biases introduced by timing of measurements and seasonal dynamics, we used this simplifying assumption to corroborate the patterns observed with the sensitivity metric.   -->

Across all individuals in the plot, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the concavity/convexity of a location. 
We used a Digital Elevation Model from a 5 m x 5 m elevation layer collected at plot establishment.
We used the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to fill and breach this layer, calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$.
We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
We estimated crown exposure for each tree at each census using crown illumination index (CII), an ordinal categorical variable from 1 to 5 (least to most exposed) measured at each dendroband census.

We calculated deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
We defined deciduousness as the mean proportion of crown leaf loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous.
Data was unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
An alternate measure of deciduousness is the duration spent at maximum crown loss.
This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for a larger list of species.  
<!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
We calculated niche breadth as the standard deviation of TWI across all observed individuals of each species in the whole plot census in 2008.  -->


***Statistical methods***

We first modelled sensitivity as a simple random intercept model

$$
sensitivity \sim 1 + (1 | Species)
$$ 

We then used the predicted sensitivities from this model and tested their correlations with deciduousness using simple linear models.

To model the interactive effects of deciduousness and wetness on sensitivity, we modelled

$$
sensitivity \sim 1 + TWI + deciduousness + TWI:decidousness
$$

We then calculated predictions across all combinations of values from observed maximum to minimum of TWI and deciduousness within this dataset.

To further disentange the microenvironmental drivers of drought sensitivity, we simultaneously modelled:

$$
sensitivity \sim DBH + CII + TWI + (DBH+CII+TWI | Species)
$$ 
and 
$$
CII \sim DBH
$$

Simultaneously modelling the effect of variables on sensitivity and the effect of DBH on CII in this way allows us to disentangle causual linkages between these processes. 
We chose to model a species random effect on all slopes here to account for variation across species instead of specific traits which may have different effects with each of the variables. 
However, we also ran simpler models with species random effects only on the intercept, the results of which are reported in the Supplementary Material. 
Across all models, we used Gaussian priors for the distribution of the response variable sensitivity, considered CII as an monotonic predictor (ordered factor) and centred and scaled the other predictors. 
We modelled CII as an ordinal categorical variable with a cumulative logit link function. 
In the combined form, the model therefore used a multivariate distribution of Gaussian and logit. 
<!-- Finally, to test full mediation effects, or the hypothesis that DBH effects on sensitivity are fully explained by its effects on exposure, we tested the same models, removing the DBH as a predictor of sensitivity. 
We then compared the fits of the partial mediation and full mediation models using Leave-One-Out Cross Validation.   -->

All statistical analyses were performed using R version 4.4.0 [@rcoreteam_language_2024]. 
Mixed models were fit through Bayesian regressions using MCMC methods using the package _brms_ [@bürkner_brms_2017].
We ran 4 MCMC chains for 3000 iterations with 1000 of these set as warmup.
We extracted draws, calculated distributions and checked convergence using functions from the _posterior_ package [@vehtari_ranknormalization_2021; @bürkner_posterior_2023].


# Results {-}

## Drought characteristics {-}

The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions.
1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought.


In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including in precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum temperature (max anomaly +1.5 SD), and VPD (max anomaly +2 SD).
1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census time period. 
Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season.
1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought.


```{r data summaries, echo=F, warning=F, message=F}
library(tidyverse)
tree.time <- readRDS("../data/HKK-dendro/sensitivity_data_formodels.RData")
# tree.time <- readRDS("data/HKK-dendro/sensitivity_data_formodels.RData")


tree.time.sp <- tree.time %>%
  group_by(spfull) %>%
  dplyr::summarise( # change to mm
    inc_annual_med = round(median(inc_annual * 10, na.rm = T), 2),
    inc_annual_sd = round(sd(inc_annual * 10, na.rm = T), 2)
  ) %>%
  arrange(inc_annual_med)

sens.sp <- tree.time %>%
  filter(yr %in% c(2010, 2015)) %>%
  group_by(spfull, yr) %>%
  dplyr::summarise(
    sens_med = round(median(sens.prop, na.rm = T), 2),
    sens_mean = round(mean(sens.prop, na.rm = T), 2),
    sens_sd = round(sd(sens.prop, na.rm = T), 2)
  ) %>%
  arrange(sens_med)

# spread this for each year

sens.sp.wide <- sens.sp %>%
  ungroup() %>%
  dplyr::select(spfull, yr, sens_med) %>%
  pivot_wider(names_from = yr, values_from = sens_med, names_prefix = "sens.")

# paired correlation of 2010 and 2015 sensitivities across species

sens.sp.cor <- cor.test(sens.sp.wide$sens.2010, sens.sp.wide$sens.2015)
```

## Growth timeseries and drought responses {-}

The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a). 
Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed. 
<!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->

Drought years were associated with lower growth across the plot. 
In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. 
The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. 
While median species responses to the 2010 drought were variable, all species had lower increment than a typical year in 2015. 
<!-- Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4).  -->
<!-- Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`.  -->
<!-- The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%. -->
<!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->

On average, trees had negative drought sensitivities for both droughts, but with considerable variation (Fig `r fig_drought_responses`b). 
Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`. 
In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`*. 
In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`*. 


```{r model summaries, echo=F, warning=F, message=F}
# load data--------------------------
datasets <- readRDS("../data/HKK-dendro/sensitivity_data.RData")

tree.time <- datasets$tree.time
tree_vars <- datasets$tree_vars
sp_vars <- datasets$sp_vars

colnames(sp_vars)[1] <- "Species"

coefs_df <- readRDS("../results/models/non_negative/sensitivity_model_intercept.RData")
ranef_df <- coefs_df %>% filter(grepl("r_Species", param))
ranef_df$Species <- gsub("r_Species\\[|\\,Intercept\\]", "", ranef_df$param)

# add intercept to ranef_df
# make intercept vector
intercepts <- coefs_df %>%
  filter(param %in% "Intercept") %>%
  dplyr::select(median) %>%
  pull(median)
intercepts <- rep(intercepts, each = nrow(ranef_df) / 2)
ranef_df <- ranef_df %>%
  dplyr::mutate(
    intercept = intercepts + median,
    lwr = intercepts + lwr,
    upr = intercepts + upr
  )

# plot intercepts against species characteristics

# join ranef_df with sp_vars
ranef_df <- merge(ranef_df, sp_vars, by = "Species", all.x = TRUE)

# lms
coefs_dec_lms <- ranef_df %>%
  # filter(yr %in% yrs) %>%
  filter(Species != "ALPHVE") %>%
  nest_by(yr) %>%
  # dplyr::mutate(mod = list(lm(intercept ~ williams_dec, data = data))) %>%
  dplyr::mutate(mod = list(cor.test(data$williams_dec, data$intercept))) %>%
  dplyr::reframe(broom::tidy(mod))


isocline_coefs <- readRDS("../results/models/non_negative/coefs_isoclines_nore.RDS")
# isocline_coefs <- readRDS("results/models/non_negative/coefs_isoclines_nore.RDS")

isocline_coefs <- do.call(rbind, isocline_coefs)
isocline_coefs_dec <- isocline_coefs %>% filter(param == "b_williams_dec")
isocline_coefs_twi <- isocline_coefs %>% filter(param == "b_twi")
isocline_coefs_inter <- isocline_coefs %>% filter(param == "b_twi:williams_dec")
# isocline_coefs_twi
# head(isocline_coefs)

coefs_spre <- readRDS("../results/models/orderedcii/coefs_rel_spre.RDS")
# coefs_spre <- readRDS("results/models/orderedcii/coefs_rel_spre.RDS")
coefs_df_spre <- do.call(rbind, coefs_spre)
coefs_spre_twi <- coefs_df_spre %>% filter(param == "b_sensprop_twi_scaled")
# head(coefs_spre_twi)
```



## Species effects {-}

Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S5), in alignment with median sensitivities across all trees. 

Deciduousness species had lower negative sensitivities in 2015, but not in 2010 (Fig `r fig_species`a). 
In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`). 
On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).  



![***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010 and 2015. a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values. Line shown for significant correlation. b) Predicted sensitivity values across the observed range of deciduousness and topographic wetness index from a model with TWI, deciduosness and their interaction across all trees modelled for each drought year separately. For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png)

## Microenvironment and individual effects {-}

Individual tree responses to drought were not correlated with each other (Fig S6). 

TWI had no significant on sensitivity in 2010 but a positive effect in 2015. 
Specifically, across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b). 

The differing effects of TWI on sensitivity in the two droughts, along with the influence of deciduousness, were confirmed by models that controlled for crown exposure and DBH as well as species variation in intercept and slopes (Fig `r fig_microenvs`a). 
These models showed small negative sensitivities in wetter sites in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a strong positive effect of TWI in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`). 
The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b). 
Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not  increase with TWI (Fig `r fig_microenvs`c). 
On the other hand, neither the mean sensitivities across species nor their response to TWI were associated with deciduousness in 2010.  

![***Figure `r fig_microenvs`: Drivers of variation in sensitivity among individuals***. All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes. a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree. b) wetness effect across all species and its relationship with deciduousness value. Line represents a significant correlation. c) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions. Black line represents overall relationship (panel a); coloured lines for each species reflects deciduousness values. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)

```{r exposure models, echo=F, warning=F, message=F}
coefs_spre_cii <- coefs_df_spre %>% filter(param == "bsp_sensprop_mocii_min1")
coefs_spre_ordcii <- coefs_df_spre %>% filter(param == "b_orderedciimin1_calcDBH_min1_scaled")
```

Crown exposure was associated with stronger negative sensitivities in both droughts.
In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 the negative effect was not as strong (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`). 
Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 (Fig `r fig_dag`b).


DBH was a significant predictor of crown exposure, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a). 
Controlling for this direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a). 


![***Figure `r fig_dag`: Effects of size, exposure and water availability*** from combined models. Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010 and b) 2015. Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size. c) Modelled sensitivity across crown exposure classes in 2010 and 2015. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png)

# Discussion {-}

<!---CONCLUSIONS---->
We show that, at least in one tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, and local environmental conditions (Table 1).
Analysing two droughts in a 14-year dendrometer band timeseries, we found lower-than-average growth in drought years, reflecting in lower plot-level growth (Fig `r fig_drought_responses`). 
Comparing across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
Despite lower absolute extremes of temperature, vapor pressure deficit, and number of dry days (Fig. 2a), drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig `r fig_species`). 
Interestingly, in 2015 (but not 2010), evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig `r fig_species`b, Fig `r fig_microenvs`b). 
While trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a). 
In other words, water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts; larger trees (with likely deeper roots) fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines in the wet-season drought (Fig `r fig_dag`). 
Taken together, these results indicate that joint and interactive effects of drought characteristics, species traits, and individual microenvironments produce enormous heterogeneity in individual and species drought responses of tropical trees.
This finding that individual droughts have divergent effects on species and individuals may contribute to the maintenance of species diveristy and ecosystem stability.   

<!----Limitations in filling gap/ comparing only two droughts---->
While our analysis of 30 species across size classes and habitats spans representative spatial and ecological variation realistically and beyond the scope of previous such analyses, we are limited in quantitatively analysing differences associated with drought characteristics because of low sample size. 
<!-- Long-timescale analyses with tree rings are limited by the fact that species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments. 
Tape measurements of growth from long-term ecological dynamics plots can be representative of species strategies but lack precision to parse signal from noise; error for each tape measure is 1mm, error for growth is at least 2mm while median annual growth across trees in HKK is ~1.4 mm and measuring deviation from this annual growth needs more precision.  -->
<!-- Traditional methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
Further, although estimates of woody growth from dendroband timeseries can have measurement biases because of stem water shrinkage and bark peeling [@chitra-tarak_it_2015], our results are robust to simplifying assumptions that account for these biases (Fig S--).
However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
The two droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological extremes and anomalies, and severity as classified the SPEI metric (Fig. `r fig_droughts`).
Thus, quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.
While our finding that a wet-season drought (2015) had greater impact on woody growth than a dry-season drought (2010) calls into question the generality of previous findings based on tree-ring chronologies from HKK [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] that tree growth is most sensitive to dry-season precipitation, care must be taken in interpretation given that the 2015 drought was more anomalous on multi-month time scales (Fig. `r fig_droughts`).
Rigorous generalization on tropical tree growth responses to any specific drought characteristic would require annual growth records from numerous species over numerous droughts, but no such data set currently exists.
Studies in temperate forests, where extensive tree-ring records are available, have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
As tropical dendrometer band records continues to grow, so will the potential to better understand how specific drought characteristics interact with species traits and microenironment to shape growth respones. 

<!-- ALT: We analysed species traits as static, but we might be missing intraspecific and plastic variation in drought sensitivity traits, as well as other axes of trait variation that might explain variation in drought sensitivity.
- Many hydraulic-related traits vary with tree height [ @vinod_thermal_2023], including the frequency of dry season deciduous leaf loss -- both within species and at the community level [@condit_ref; @meakem].  *caveat for species-level traits*  
- While leaf deciduousness was the main axis of hydraulic conservative strategy that we considered, leaf-level stomatal regulation and belowground root loss etc. are other potential strategies for conservative water use, and our interpretation of species patterns may be limited by our understanding of these processes across species.  
- Surprisingly, although deciduousness may have had an advantage in the dry season drought, because this strategy -> conservative water use in the dry season, species differences in this drought were not predicted by deciduousness.   -->

<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
We have demonstrated that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry tropical forest.
Tropical dry forests represent periodic hot/dry conditions, creating selective pressures for adaptative variation towards drought resistance, and this might mean that species in tropical dry forests exhibit a greater diversity of drought responess than species in other bioclimatic zones where adaptive pressures and thus ecological variability are different. 
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies and has resulted in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022), but it may not apply to the wet or everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironment.
Again, continued expansion of dendrometer band records and ecophysiological study of tropical tree drought responses will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   -->

<!-- ALT: Differential allocation to growth may not mean negative outcomes, it might mean demographic trade-offs
- Differences in allocation to growth may not be the complete picture - may have altered allocation to mortality or reproduction.  
- Some evidence that both understory and emergent species can have reduced survival during drought [@machado_forest_2023].  *for where else might they be allocating to?*
- 2015 ENSO drought changed regeneration dynamics in Thailand [@nutiprapun_effects_2023]
- also NSCs and then compensatory growth
- deciduous species have higher survival under multi-year droughts [@aguirre-guiterrez refs] -->

<!-- contributions / disentangling exposure from DBH-->
One variable examined here -- crown exposure -- had a consistent effect across the two droughts (Table 1, Figs. `r fig_microenvs`, `r fig_dag`). 
Trees with more exposed crowns exhibited greater growth declines in both droughts, while sub-canopy trees grew more than average during the 2010 drought.
This finding is consistent with the concept that larger trees tend to suffer more during drought [@bennett_larger_2015], but adds the important insight that it is crown position -- as opposed to size *per se* -- that is primarily responsible for this pattern.
The strong and consistent negative effect of crown position makes sense in light of pronounced gradients in drought stress across forest vertical profiles [@vinod_thermal_2023].
We show that residual effects of size, presumably including root water access and biophysical challenges linked to height itself, may act in different directions depending on the drought characteristics (Figs. `r fig_microenvs`, `r fig_dag`).
Together, these findings imply that while canopy and emergent trees face the greatest stress and exhibit the most negative growth responses in almost any drought, a suite of species traits and microenvironmental characteristics modify growth responses in interaction with drought characteristics.


<!----Contributions beyond---->
Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across droughts. 
An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the two events; it's growth nearly doubled in the dry-season drought but was severely reduced in the wet-season drought (Fig. S5).
More generally, tree-ring records are likely to give a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to directly and interactively influence drought sensitivity.<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. -->
Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies make particular species and trees resistant to some droughts but vulnerable to others. 
These distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought, in alignment with recent global analyses (@liu_drought_2022).    

<!----Science is better now---->
Our study reveals that drought sensitivity is not a static trait associated to particular species, traits, or microenvironments, but rather a dynamic response shaped interactively by these factors and drought characteristics. 
Despite net decreased growth at the plot level (in agreement with landscape-scale in @aguirre-gutiérrez_functional_2022 and @bennett_sensitivity_2023), we show evidence that the same strategies are not uniformly successfully across spatiotemporal climatic variation. 
Static traits are often used to explain and predict forest responses to drought, especially in the tropics where long-term records are few [@guillemot_small_2022; @vico_how_2017a]. 
<!--aligns with remote sensing based inferences across global forests @liu_drought_2022-->
More inclusive frameworks for drought responses encompass ecological complexity and covarying traits with environment [e.g., @trugman_why_2021] but still assume consistent sensitivity of individual trees across events.
However, we demonstrate that drought responses of tropical trees are not linearly predictable across drought events, but rather are shaped by the unique characteristics of each drought based on interactions between species adaptations and local environmental conditions.
Without overriding the net negative effects of drought on tropical tree growth, these complexities reduce the likelihood of any given species or tree suffering severe growth declines under multiple consecutive droughts, thereby promoting diversity and stabilizing functioning of diverse tropical forests under drought.  


<!---Other points--->

<!-- - (BCI light x size interaction in @rüger_growth_2011)   -->



# Acknowledgements {-}

We acknowledge the continued contributions of the Huai Kha Khaeng forest monitoring team in collecting data and maintaining the dendrometer bands. 
Ekaphan Kraichak and Kanisorn Chowtiwuttakorn helped with procuring elevation and climate datasets. 
José Medina-Vega, Valentine Herrmann and Eugenie Más provided input on improving analyses. 
KA was supported by the Smithsonian Institution Postdoctoral Fellowship and the ForestGEO Analytical Workshop, Asia 2024 in completing this work. 

\newpage
# References  {-}

