---
title: "Climate Sensitivity of Tropical Tree Growth"
author: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## HKK climate

```{r echo = FALSE, message=F, warning=F}

#load libraries
library(tidyverse)
library(lubridate)
library(httr)
library(janitor)

#reading the GitHub PAT (stored privately and locally)
pat<-as.character(read.table("../data/HKK-dendro/git_pat.txt")[1,1])

# remotes:::download("DendroByBand.RData",
#                    "https://raw.githubusercontent.com/forestgeo/HKK-dendrobands/main/data/modified_data/DendroByBand.RData",
#                    auth_token = pat)

load("../data/HKK-dendro/DendroByBand.RData")

dendrobyband<-merge(dendrobyband, trees, by="Tag")

hkk_OBSID$year<-year(hkk_OBSID$Date2)
hkk_OBSID$month<-month(hkk_OBSID$Date2)
hkk_OBSID$day<-day(hkk_OBSID$Date2)

#correcting two sets of wrong dates

hkk_OBSID$year[hkk_OBSID$year==2209]<-2009
hkk_OBSID$year[hkk_OBSID$month==1 & hkk_OBSID$Cno==5]<-2011

dendrobyband<-merge(dendrobyband, hkk_OBSID, by=c("OBSID", "Cno", "Date2", "Tag"))

dendrobyband$Date2<-as.Date(paste0(dendrobyband$year,"-", dendrobyband$month, "-", dendrobyband$day))

#this can be refined if needed
wet_census<-c(1, 3, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28)


#getting climate data for all


#CHIRPS data ----

# CHIRPS data GEE code here - https://code.earthengine.google.com/894597bd093906fb0ff82320f29c0842


chirps_hkk<-read.csv("../data/climate/CHIRPS_from_GEE.csv")
colnames(chirps_hkk)[1]<-"time"

#making dates manageable
chirps_hkk<- chirps_hkk %>%
  dplyr::mutate(time=as.Date(time, format = "%b %d, %Y"),
                year=year(time),
                month=month(time))

census_int<-dendrobyband%>%
  group_by(Cno) %>%
  dplyr::summarise(start=min(Date2, na.rm = T),
                   end=max(Date2, na.rm=T))%>%
  dplyr::mutate(type=ifelse(Cno %in% wet_census, "wet", "dry"))

pre_yr_census<-ggplot()+
  geom_area(data=chirps_hkk, aes(x=time, y=precipitation), fill="blue", alpha=0.5)+
  #geom_line(data=climate_data_orig[climate_data_orig$variable=="tmx",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value*5), col="red")+
  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=start), col="darkgreen")+
  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=end), col="darkgreen")+
  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=start), col="brown4")+
  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=end), col="brown4")+
  # scale_y_continuous(    # Features of the first axis
  #   name = "monthly rainfall (mm)",
  #   
  #   # Add a second axis and specify its features
  #   sec.axis = sec_axis(~.*0.2, name="maximum temperature (Celsius °)"))+
  xlab("Date")+theme_bw()+
  ggtitle("HKK dendroband censuses and CHIRPS rainfall data")+
  #scale_x_continuous(n.breaks=15)+
  theme(axis.text.x = element_text(angle=45))

 
pre_yr_census

#rainfall per year

#-----------------Decisions on climate data-----------------------=
# Summarising climate data to relevant time window for SCBI
# As per issue #42 and KAT pub https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.15934
# HKK best climate variables are:
# PPT - previous Sep to current June (it is actually precipitation day freq, but using sum now)
# Tmax - current Apr to Oct
#---------------------------------------------------------------=

#applying these decisions and rescaling

#mean annual precip from 1981 to 2023 is 1698.4

#code - https://code.earthengine.google.com/b0eda787f2e1ba578bbf50a8a9d04843

climate_data_yr<-chirps_hkk %>%
  group_by(year) %>%
  summarise(
    tot_precip=sum(precipitation),
    pre_SeptoDec=sum(precipitation[month>=9]), #Sep to Dec precip for each year
    pre_JantoJun=sum(precipitation[month<=6]) #Jan to June precip for each year
    #tmx_window=sum(value[variable == "tmx"& month %in% 4:10])
)%>% #sum of potential evapotranspiration for current May to July
  mutate(
    pre_window=pre_JantoJun+lag(pre_SeptoDec), #adding prev year's June to Dec and current year's Jan to June
    tot_pre_diff=tot_precip-1698.4
    )

#standard deviation is 254.06

precip_plot<-ggplot(climate_data_yr)+
  geom_bar(aes(x=year, y=tot_pre_diff), stat="identity")+
  ylab("difference from 40-year mean (mm)")+
  geom_hline(yintercept = c(254.06, -254.06), lty=2)+
  theme_bw()
precip_plot




# library
library(ggridges)
library(ggplot2)
library(viridis)
library(hrbrthemes)

chirps_hkk$year_fac<-as.character(chirps_hkk$year)
chirps_hkk$year_fac<-factor(chirps_hkk$year)


# Plot
ggplot(chirps_hkk, aes(x = precipitation, y = year_fac, 
                       fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "rainfall", option = "C") +
  labs(title = 'Rainfall in HKK') +
  #theme_ipsum() +
  theme_bw()+
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )



```
```{r echo=F}

#remove all data with decimal errors, misplaced decimal

dendro.data<-dendrobyband %>%
  dplyr::rename(sp=SPCODE.UPDATE)%>%
  mutate(year=year(Date2)) %>%
  #filter(year!=2209)%>% #quick fix to remove bad dates - there are 30 entries
  #filter(start_date <= year & year <= end_date) %>%
  filter(Cno %in% wet_census)%>%
  filter(is.na(DBH1_flag), DBH1_error=="none",
         win.decimal.error=="none", is.na(win.flag))%>%
  group_by(year, sp, Tag) %>%
  dplyr::summarise(calcDBH=calcDBH[which(n.dendro==min(n.dendro))],
                   Date2=Date2[which(n.dendro==min(n.dendro))],
                   Cii=max(Cii, na.rm=T))



dendro.data_all<-dendro.data%>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=year-lag(year),
         #removing increment value if the previous measure was before the last year
         inc_annual=ifelse(inc.time==1, inc_annual, NA),
         #has.out=ifelse(1 %in% out, 1, 0),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  ungroup()%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  #filter(start_date <= year & year <= end_date) %>%
  group_by(sp, treeID, year)%>%
  dplyr::summarise(
    diainc = mean(inc_annual, na.rm=T),
    Cii=max(Cii, na.rm=T))

#first fitting a line through the time series
obs.model.dbh<-nlme::lme(calcDBH~year, random=~1|Tag, data=dendro.data,
                         na.action=na.omit,
                         method="ML", control=(msMaxIter=100)) #had to increase the number of iterations

#standardised residuals calculated https://stat.ethz.ch/R-manual/R-devel/library/nlme/html/residuals.lme.html

std.res.dbh<-residuals(obs.model.dbh, type="pearson")
hist(std.res.dbh)
dendro.data$std.res<-rep(NA, nrow(dendro.data))
dendro.data$std.res[!is.na(dendro.data$calcDBH)]<-std.res.dbh
#defining all observation with abs(residual) >3 as an outlier
dendro.data$out<-factor(ifelse(abs(dendro.data$std.res)<=3, 0, 1))

#find the trees which have outlier observations
outs.Tag<-unique(dendro.data$Tag[dendro.data$out==1])

#plot trees that have outliers
ggplot(dendro.data[dendro.data$Tag %in%outs.Tag,],
       aes(x = year, y = calcDBH, col=factor(out), group = Tag))+
  geom_line(alpha=0.5)+geom_point()+
  scale_color_manual(values=c("grey20", "red", "grey"))

dendro.data.noouts<-dendro.data %>%
  mutate(dbh3=ifelse(abs(std.res)<=3, calcDBH, NA)) %>% #create an outlier removed column
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=year-lag(year),
         #removing increment value if the previous measure was before the last year
         #inc=ifelse(inc.time==1, inc, NA),
         inc_annual=ifelse(inc.time==1, inc_annual, NA),
         has.out=ifelse(1 %in% out, 1, 0),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  ungroup()%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  #filter(start_date <= year & year <= end_date) %>%
  group_by(sp, treeID, year)%>%
  dplyr::summarise(
    diainc = mean(inc_annual, na.rm=T),
    calcDBH_min1 = mean(calcDBH_min1, na.rm=T),
    dbh=mean(calcDBH, na.rm=T),
    out = ifelse(1 %in% out, 1, 0),
    has.out = ifelse(1 %in% has.out, 1, 0), 
    cii=max(Cii, na.rm=T))%>%
  filter(out==0)%>%
  ungroup() %>%
  dplyr::mutate(
    diainc.scaled = scale(diainc),
    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
  filter(out.inc==0)

# species_all<-dendro.data.noouts %>%
#   group_by(sp, treeID) %>%
#   dplyr::summarise(n.obs=length(treeID))%>%
#   ungroup() %>%
#   group_by(sp) %>%
#   dplyr::summarise(n.tree=length(treeID),
#                    n.obs.tot=sum(n.obs))%>%
#   ungroup()%>%
#   arrange(-n.tree)
# 
# write.csv(species_all, "../data/HKK-dendro/species.csv", row.names = F)

```

## Raw diameter increments of all species

```{r echo=FALSE, eval=F}

library(ggplot2)

growth_yr_withouts<-ggplot(dendro.data_all, aes(x=year, y=diainc, group=year))+
  geom_violin()+
  ylab("Annualised diameter increment (cm)")+theme_bw()

growth_yr_withouts
```

## Removing outliers based (more than 3 SD)  

```{r echo=FALSE}


growth_yr_noouts<-ggplot(dendro.data.noouts, aes(x=year, y=diainc, group=year))+ geom_violin()+ ylab("Annualised diameter increment (cm)")+theme_bw()

growth_yr_noouts
```


```{r maps of species with tree ring records, echo=F, eval=F}
library(ggplot2)

pat<-as.character(read.table("git_pat.txt")[1,1])

remotes:::download("stem1.RData",
                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
                   auth_token = pat)
remotes:::download("stem2.RData",
                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
                   auth_token = pat)
remotes:::download("stem3.RData",
                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
                   auth_token = pat)
remotes:::download("stem4.RData",
                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
                   auth_token = pat)


load("stem1.RData")
load("stem2.RData")
load("stem3.RData")
load("stem4.RData")

hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)

hkk_census <- hkk_census %>%
  filter(status=="A")

hkk_census_ring_sp<-hkk_census %>%
  filter(CensusID==3)%>%
  filter(sp %in% ring.sp)

hkk_census_sp <- hkk_census %>%
  filter(CensusID==4)%>%
  #group_by(sp, tag, CensusID) %>%
  #dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
  group_by(sp, tag)%>%
  dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
  group_by(sp) %>%
  dplyr::summarise(n.tree=length(unique(tag)),
                   sum.agb=sum(tot.agb))%>%
  ungroup()%>%
  dplyr::mutate(pct.agb=sum.agb*100/sum(sum.agb, na.rm=T))%>%
  arrange(-pct.agb)%>%
  dplyr::mutate(pct.agb.cum=cumsum(pct.agb))

sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA",
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")

#pct agb of the top 15 dendro species
sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])

```


## Detrending all species

```{r detrending all, echo=F}

dendro.data.noouts<-dendro.data.noouts %>%
  group_by(sp, treeID) %>%
  dplyr::mutate(n.obs=length(treeID))%>%
  filter(n.obs>=4)%>%
  ungroup()%>%
  filter(treeID!="HKK.92562")%>% #removing an issue with spline for now
  group_by(sp, treeID)%>%
  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))
```

For `r length(unique(dendro.data.noouts$sp))` species with dendrobands, all individual trees detrended for DBH with smooth splines

```{r echo=F}

growth_yr_detrend_all<-ggplot(dendro.data.noouts, aes(x=year, y=inc.resid, group=year))+ 
  #geom_violin()+ 
  geom_boxplot()+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))
growth_yr_detrend_all


```

## top 10 species

```{r echo=F}

dendro.by.sp<-dendro.data.noouts %>%
  group_by (sp, year) %>%
  dplyr::summarise(n.inds=length(sp),
    mean.inc = mean(inc.resid, na.rm=T),
                   median.inc = median(inc.resid, na.rm=T),
                   sd.inc = sd(inc.resid, na.rm=T),
    lower=mean.inc-sd.inc,
    upper=mean.inc+sd.inc)%>%
  filter(n.inds>=10)%>%
  arrange(year, -median.inc)


sp_top5<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA")

sp_top10<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA") 


dendro.top5<-dendro.by.sp %>%
  filter(sp %in% sp_top5)


dendro.top5.dat<-dendro.data.noouts %>%
  filter(sp %in% sp_top5)

dendro.top10.dat<-dendro.data.noouts %>%
  filter(sp %in% sp_top10)

hkk_census_top10_sp<-hkk_census %>%
  filter(CensusID==3)%>%
  filter(sp %in% sp_top10)

#pct agb of the top 15 dendro species
sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])

top10sp.map.census<-ggplot(hkk_census_top10_sp, aes(x=gx, y=gy, 
                                                    #size=as.numeric(dbh/10),
                                                    color=factor(sp, levels=sp_top10)))+
  geom_point(size=1)+
   scale_size(range = c(.1, 10), name="dbh")+
  guides(color=guide_legend(title="Species"))+
  scale_color_viridis_d()+
  ggtitle("top 10 species in 2004")+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

top10sp.map.census

trees.top10<-trees[trees$SPCODE.UPDATE %in% sp_top10,]


top10sp.map.dendro<-ggplot(trees.top10, aes(x=X, y=Y, 
                                                    #size=as.numeric(dbh/10), 
                                                    color=factor(SPCODE.UPDATE, levels=sp_top10)))+
  geom_point(size=1)+
   scale_size(range = c(.1, 10), name="dbh")+
  guides(color=guide_legend(title="Species"))+
  scale_color_viridis_d()+
  ggtitle("top 10 species in 2004")+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

top10sp.map.dendro

```
## Fastest and slowest in 2015 and 2016

```{r echo=F}

dendro.data.noouts <- dendro.data.noouts %>%
  group_by(year) %>% 
  dplyr::mutate(growth_quantile = ntile(diainc, 4),
                resid_quantile = ntile (inc.resid, 4))
  
growth_2015 <- dendro.data.noouts %>%
  filter(year==2015)

quantiles_2015<-quantile(growth_2015$diainc, probs=c(0.25, 0.5, 0.75, 1))

growth_2015_plot<-ggplot(growth_2015, aes(x=diainc)) +
  geom_density(fill="brown", col="brown", alpha=0.5) +
  geom_vline(xintercept=quantiles_2015, col="brown")+ theme_bw()
  

growth_2014 <- dendro.data.noouts %>%
  filter(year==2014)

quantiles_2014<-quantile(growth_2014$diainc, probs=c(0.25, 0.5, 0.75, 1))

growth_2016 <- dendro.data.noouts %>%
  filter(year==2016)

quantiles_2016<-quantile(growth_2016$diainc, probs=c(0.25, 0.5, 0.75, 1))

growth_2016_plot<-ggplot(growth_2016, aes(x=diainc)) +
  geom_density(fill="brown", col="brown", alpha=0.5) +
  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()

growth_2017 <- dendro.data.noouts %>%
  filter(year==2017)

quantiles_2017<-quantile(growth_2017$diainc, probs=c(0.25, 0.5, 0.75, 1))

growth_2017_plot<-ggplot(growth_2017, aes(x=diainc)) +
  geom_density(fill="brown", col="brown", alpha=0.5) +
  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()

growth_2014_2015_plot<-ggplot() +
  geom_density(data = growth_2014, aes(x=diainc), fill="forestgreen", col="forestgreen", alpha=0.5) +
  geom_vline(xintercept=quantiles_2014, col="forestgreen")+
  geom_density(data= growth_2015, aes (x=diainc), fill="brown", col="brown", alpha=0.5) +
  geom_text(aes(x=0.75, y=1, label="2014"), col="forestgreen", size=3)+
  geom_text(aes(x=0.75, y=1.5, label="2015"), col="brown", size=3)+
  geom_vline(xintercept=quantiles_2015, col="brown")+ 
  xlab("diameter increment (cm)") + ylab("density")+
  theme_bw()
  

dendro.data.noouts$year_fac<-as.character(dendro.data.noouts$year)
dendro.data.noouts$year_fac<-factor(dendro.data.noouts$year_fac, levels=as.character(rev(2009:2023)))

growth_all_years<-ggplot(dendro.data.noouts, aes(x = diainc, y = year_fac, 
                       fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "growth", option = "H", direction=-1) +
  labs(title = 'tree growth in HKK (cm)', y="year") +
  #theme_ipsum() +
  geom_vline(xintercept=0, lty=2)+
  theme_bw()+
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )

growth_all_years

png("../doc/display/growth_2014_2015.png", width=4, height=4, units="in", res=300)
growth_2014_2015_plot
dev.off()

png("../doc/display/growth_allyrs.png", width=4, height=4, units="in", res=300)
growth_all_years
dev.off()


```

##Distribution of data with size, exposure and deciduousness

```{r echo=F}

#2014 data distribution with size, exposure and deciduousness

png("../doc/display/dbh_cii.png", width=4, height=4, units="in", res=300)
ggplot(growth_2014, aes(x = dbh, y = factor(cii), 
                       fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
  labs(title = 'DBH in 2014 (cm)', y="canopy illumination") +
  #theme_ipsum() +
  theme_bw()+
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )

length(unique(growth_2014$sp))

png("../doc/display/growth_cii.png", width=4, height=4, units="in", res=300)
ggplot(growth_2014, aes(x = diainc, y = factor(cii), 
                       fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
  labs(title = 'growth in 2014 (cm)', y="canopy illumination") +
  #theme_ipsum() +
  theme_bw()+
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )
dev.off()

```


## Growth of most abundant species

```{r echo=F}

growth_sp5<-ggplot(dendro.top5.dat, aes(x=factor(sp, levels=sp_top5), y=diainc,
                                              fill=factor(sp, levels=sp_top5),
                                              color=factor(sp, levels=sp_top5)))+ 
  geom_violin(position = "dodge")+ 
    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top5.dat$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.top5.dat$diainc, na.rm=T), lty=2)+
  geom_text(aes(label=paste0("median=\n", round(median(dendro.top5.dat$diainc, na.rm=T),2), " cm"), 
                x=0.6, y=median(dendro.top5.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+
  xlab("")

growth_sp5

growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              fill=factor(sp, levels=sp_top10),
                                              color=factor(sp, levels=sp_top10)))+ 
  geom_violin(position = "dodge")+ 
    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_text(aes(label=paste0("median=\n", round(median(dendro.top10.dat$diainc, na.rm=T),2), " cm"), 
                x=0.6, y=median(dendro.top10.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+
  xlab("")

growth_sp10


growth_yr_sp<-ggplot(dendro.top5, aes(x=year, y=mean.inc, 
                                      color=sp))+ 
  geom_point()+
  geom_errorbar(aes(x=year, ymin=lower, ymax=upper))+
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))+
  facet_grid(sp~.)+
  guides(color="none")+
  geom_hline(yintercept = 0)+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  geom_vline(xintercept=c(2010, 2016), lty=2)
growth_yr_sp

#png("plots/growth_by_sp.png", height=8, width=4, units="in", res=300)
#growth_yr_sp
#dev.off()

```

## Size classes

```{r echo=F}

size_top10<-ggplot(dendro.top10.dat, aes(x=dbh))+
  geom_density()+
  geom_vline(aes(xintercept=35))+
  geom_vline(aes(xintercept=75))

size_top10

#trying three categories
# <50, 50-100, >100

dendro.top10.dat<-dendro.top10.dat %>%
  dplyr::mutate(size=ifelse(dbh<35, "< 35 cm dbh",
                       ifelse(dbh<70, "35-75 cm dbh", ">75 cm dbh")))


growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
      theme(axis.text.y = element_blank())+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+coord_flip()

growth_sp10

growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh")), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+

  geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge", col="black", fill="white", width=0.1)+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  #facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
      theme(axis.text.x = element_text(angle=45))+
  xlab("")+
  facet_grid(.~factor(sp, levels=sp_top10))+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))#+coord_flip()

growth_sp10

```

## Canopy illumination

```{r echo=F}

cii_top10<-ggplot(dendro.top10.dat, aes(x=cii))+
  geom_histogram()

cii_top10

#trying three categories
# <50, 50-100, >100

#dendro.top10.dat<-dendro.top10.dat %>%
#  dplyr::mutate(size=ifelse(dbh<50, "small",
#                       ifelse(dbh<100, "medium", "large")))


growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
                        aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(cii~.)+
      theme(axis.text.y = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+coord_flip()

growth_sp10_cii

growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
                        aes(
                          #x=factor(sp, levels=sp_top10), 
                          x=factor(cii), 
                            y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(.~factor(sp, levels = sp_top10))+
      theme(axis.text.y = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  #coord_flip()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))
growth_sp10_cii

```
## spatial patterns

```{r spatial patterns, echo=F}

trees.top10 <- trees.top10 %>%
  dplyr::mutate(treeID=paste0("HKK.", Tag),
                plotID=paste0(Col, "_", Row))

dendro.top10.dat<-merge(dendro.top10.dat, 
                        dplyr::select(trees.top10, c("Col", "Row", "treeID", "X", "Y", "plotID")), 
                        by="treeID", all.x=T)

dendro.top10.dat<-dendro.top10.dat %>%
  dplyr::mutate(plotID=paste0(Col, "_", Row))
  

census_spatial<-hkk_census %>%
  group_by(quadrat) %>%
  dplyr::summarise(sp.no=length(unique(sp)),
                   sp.top.10 = length(unique(sp %in% sp_top10)))%>%
  dplyr::mutate(Col=as.numeric(str_sub(quadrat, 1, 2)),
                Row=as.numeric(str_sub(quadrat, 3, 4)))

tiles<-ggplot(census_spatial, aes(x=Col, y=Row, fill=sp.no))+
  geom_tile()+
  #facet_wrap(~year)+
  coord_fixed()+
  #xlim(c(0,49))+ylim(c(0, 24))+
  theme_bw()

tiles

tiles<-ggplot(census_spatial, aes(x=Col, y=Row, fill=sp.top.10))+
  geom_tile()+
  #facet_wrap(~year)+
  coord_fixed()+
  #xlim(c(0,49))+ylim(c(0, 24))+
  theme_bw()

tiles

spatial_summary<- dendro.top10.dat %>%
  group_by(Col, Row, plotID, sp, year) %>%
  dplyr::summarise(inc.sum=sum(inc.resid, na.rm=T)) %>%
  group_by(Col, Row, plotID, year) %>%
  dplyr::summarise(inc.diff=var(inc.sum, na.rm=T))

  
tiles<-ggplot(spatial_summary, aes(x=Col, y=Row, fill=log(inc.diff, 10)))+
  geom_tile()+
  facet_wrap(~year)+coord_fixed()+
  xlim(c(0,49))+ylim(c(0, 24))+
  theme_bw()

tiles

```

## evergreen/dec

```{r evg/dec, echo=F}

sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA") 

#these 15 species had the most dendrobands and together made up 51.22% of the agb in 2009

#below values eyeballed from Williams et al. 2008 Fig 1.
sp_top15_phenoscore<-c(2.5, 0.8, 1.3, 4, 0.7, 
                       1, 1.25, 1.27, 0.2, 0.3,
                       0.4, 0.65, 3.95, 0.15, 0.25)

#no info on Alphonsea ventricosa in the paper. 
#Apparently evergreen, so used 1 as value - middle of evergreen range


pheno<-as.data.frame(cbind(sp_top15, sp_top15_phenoscore))
colnames(pheno)<-c("sp", "phenoscore")

dendro.top10.dat<-merge(dendro.top10.dat, pheno, by="sp")
  
#categorising as evg and deciduous
dendro.top10.dat$cat<-ifelse(dendro.top10.dat$phenoscore<=1, "evg", "dec")

growth_yr_cat_detrend<-ggplot(dendro_top15, aes(x=factor(cat), y=inc.resid,
                                              #group=c(year, sp),
                                              #group=sp,
                                              fill=factor(cat)))+ 
  geom_violin(position = "dodge")+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  facet_grid(.~year)+
  geom_hline(yintercept = 0)+
  scale_fill_manual(values=c("sienna4", "darkgreen"))+
  theme(axis.text.x = element_blank())

growth_yr_cat_detrend


growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
                        aes(
                          #x=factor(sp, levels=sp_top10), 
                          x=factor(cii), 
                            y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(cat~factor(sp, levels = sp_top10))+
      theme(axis.text.y = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  #coord_flip()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))
growth_sp10_cii

```


## Repeat all steps for dry censuses

```{r echo=F}

#remove all data with decimal errors, misplaced decimal

`%nin%` <- Negate(`%in%`)

dendro.data<-dendrobyband %>%
  dplyr::rename(sp=SPCODE.UPDATE)%>%
  mutate(year=year(Date2)) %>%
  #filter(year!=2209)%>% #quick fix to remove bad dates - there are 30 entries
  #filter(start_date <= year & year <= end_date) %>%
  filter(Cno %nin% wet_census)%>%
  filter(is.na(DBH1_flag), DBH1_error=="none",
         win.decimal.error=="none", is.na(win.flag))%>%
  group_by(year, sp, Tag) %>%
  dplyr::summarise(calcDBH=calcDBH[which(n.dendro==min(n.dendro))],
                   Date2=Date2[which(n.dendro==min(n.dendro))],
                   Cii=max(Cii, na.rm=T))



dendro.data_all<-dendro.data%>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=year-lag(year),
         #removing increment value if the previous measure was before the last year
         inc_annual=ifelse(inc.time==1, inc_annual, NA),
         #has.out=ifelse(1 %in% out, 1, 0),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  ungroup()%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  #filter(start_date <= year & year <= end_date) %>%
  group_by(sp, treeID, year)%>%
  dplyr::summarise(
    diainc = mean(inc_annual, na.rm=T),
    Cii=max(Cii, na.rm=T))

#first fitting a line through the time series
obs.model.dbh<-nlme::lme(calcDBH~year, random=~1|Tag, data=dendro.data,
                         na.action=na.omit,
                         method="ML", control=(msMaxIter=100)) #had to increase the number of iterations

#standardised residuals calculated https://stat.ethz.ch/R-manual/R-devel/library/nlme/html/residuals.lme.html

std.res.dbh<-residuals(obs.model.dbh, type="pearson")
hist(std.res.dbh)
dendro.data$std.res<-rep(NA, nrow(dendro.data))
dendro.data$std.res[!is.na(dendro.data$calcDBH)]<-std.res.dbh
#defining all observation with abs(residual) >3 as an outlier
dendro.data$out<-factor(ifelse(abs(dendro.data$std.res)<=3, 0, 1))

#find the trees which have outlier observations
outs.Tag<-unique(dendro.data$Tag[dendro.data$out==1])

#plot trees that have outliers
ggplot(dendro.data[dendro.data$Tag %in%outs.Tag,],
       aes(x = year, y = calcDBH, col=factor(out), group = Tag))+
  geom_line(alpha=0.5)+geom_point()+
  scale_color_manual(values=c("grey20", "red", "grey"))

dendro.data.noouts<-dendro.data %>%
  mutate(dbh3=ifelse(abs(std.res)<=3, calcDBH, NA)) %>% #create an outlier removed column
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=year-lag(year),
         #removing increment value if the previous measure was before the last year
         #inc=ifelse(inc.time==1, inc, NA),
         inc_annual=ifelse(inc.time==1, inc_annual, NA),
         has.out=ifelse(1 %in% out, 1, 0),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  ungroup()%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  #filter(start_date <= year & year <= end_date) %>%
  group_by(sp, treeID, year)%>%
  dplyr::summarise(
    diainc = mean(inc_annual, na.rm=T),
    calcDBH_min1 = mean(calcDBH_min1, na.rm=T),
    dbh=mean(calcDBH, na.rm=T),
    out = ifelse(1 %in% out, 1, 0),
    has.out = ifelse(1 %in% has.out, 1, 0), 
    cii=max(Cii, na.rm=T))%>%
  filter(out==0)%>%
  ungroup() %>%
  dplyr::mutate(
    diainc.scaled = scale(diainc),
    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
  filter(out.inc==0)


```

## Raw diameter increments of all species

```{r echo=FALSE}

library(ggplot2)

growth_yr_withouts<-ggplot(dendro.data_all, aes(x=year, y=diainc, group=year))+
  geom_violin()+
  ylab("Annualised diameter increment (cm)")+theme_bw()

growth_yr_withouts
```

## Removing outliers based (more than 3 SD)  

```{r echo=FALSE}


growth_yr_noouts<-ggplot(dendro.data.noouts, aes(x=year, y=diainc, group=year))+ geom_violin()+ ylab("Annualised diameter increment (cm)")+theme_bw()

growth_yr_noouts
```

## Tree ring species

- Four species with tree ring chronologies in HKK

```{r ring trees, echo=F}
#These are species that have been cored near HKK and reported in Zuidema et al., Anderson-Teixeira et al., etc
ring.sp<-c("AFZEXY", "MELIAZ", "TOONCI", "CHUKTA", "NEOLOB")

trees.with.ring<-trees[trees$SPCODE.UPDATE %in% ring.sp,]

dendro.ring<-dendro.data.noouts %>%
  filter(sp %in% ring.sp)

hkk_sp_ring<-trees.with.ring %>%
  group_by(SPCODE.UPDATE)%>%
  dplyr::summarise(n.tree=length(SPCODE.UPDATE))
knitr::kable(hkk_sp_ring, format='html', table.attr="style='width:50%;'")

```

```{r maps of species with tree ring records, echo=F}
library(ggplot2)

pat<-as.character(read.table("git_pat.txt")[1,1])

# remotes:::download("stem1.RData",
#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
#                    auth_token = pat)
# remotes:::download("stem2.RData",
#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
#                    auth_token = pat)
# remotes:::download("stem3.RData",
#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
#                    auth_token = pat)
# remotes:::download("stem4.RData",
#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
#                    auth_token = pat)
# 

load("stem1.RData")
load("stem2.RData")
load("stem3.RData")
load("stem4.RData")

hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)

hkk_census <- hkk_census %>%
  filter(status=="A")

hkk_census_ring_sp<-hkk_census %>%
  filter(CensusID==3)%>%
  filter(sp %in% ring.sp)

hkk_census_sp <- hkk_census %>%
  filter(CensusID==4)%>%
  #group_by(sp, tag, CensusID) %>%
  #dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
  group_by(sp, tag)%>%
  dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
  group_by(sp) %>%
  dplyr::summarise(n.tree=length(unique(tag)),
                   sum.agb=sum(tot.agb))%>%
  ungroup()%>%
  dplyr::mutate(pct.agb=sum.agb*100/sum(sum.agb, na.rm=T))%>%
  arrange(-pct.agb)%>%
  dplyr::mutate(pct.agb.cum=cumsum(pct.agb))

sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA",
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")

#pct agb of the top 15 dendro species
sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])

ring.sp.map.census<-ggplot(hkk_census_ring_sp, aes(x=gx, y=gy, size=as.numeric(dbh/10), color=factor(sp)))+
  geom_point(alpha=0.5)+
   scale_size(range = c(.1, 10), name="DBH (cm)")+
  guides(color=guide_legend(title="Species"))+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

ring.sp.map.census

ring.sp.map<-ggplot(trees.with.ring, aes(x=X, y=Y, size=as.numeric(DBH.04), color=factor(SPCODE.UPDATE)))+
  geom_point(alpha=0.5)+
   scale_size(range = c(.1, 10), name="DBH (cm)")+
  guides(color=guide_legend(title="Species"))+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

ring.sp.map
```
Species with tree rings contribute `r round(sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% ring.sp]),2)` % aboveground biomass in 2004

## Dendrobands on ring species

```{r echo=F}

growth_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
                                              fill=factor(sp),
                                              color=factor(sp)))+ 
  geom_violin(position = "dodge")+ 
  geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #facet_grid(.~year)+
  #geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  geom_text(aes(label=paste0("median=\n", round(median(dendro.ring$diainc, na.rm=T),2), " cm"), 
                x=0.6, y=median(dendro.ring$diainc, na.rm=T)+0.15), col="black", size=3)+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))

growth_sp

growth_sp_box<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
                                              #color=factor(sp),
                                              fill=factor(sp)))+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #facet_grid(.~year)+
  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))

#growth_sp_box

growth_yr_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
                                              #color=factor(sp),
                                              fill=factor(sp)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position="dodge")+
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  facet_grid(.~year)+
  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
  theme(axis.text.x = element_blank())+
  guides(color="none")

growth_yr_sp

```

## Detrending by tree with smooth splines

```{r detrending, echo=F}

#trying for one tree

dendro.ring<-dendro.ring %>%
  group_by(sp, treeID) %>%
  dplyr::mutate(n.obs=length(treeID))%>%
  filter(n.obs>=4)%>%
  ungroup()%>%
  group_by(sp, treeID)%>%
  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))

growth_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), 
                                           y=inc.resid, 
                                           #color=factor(sp),
                                           fill=factor(sp)))+ 
  #geom_violin()+ 
  geom_boxplot()+ 
  ylab("Residuals of annualised diameter increment")+
  theme_bw()+
  theme(axis.text.x = element_text(angle=45))
growth_sp_detrend

growth_yr_detrend<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=year))+ geom_boxplot()+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))
growth_yr_detrend

```

## Detrended plots by species

```{r echo=F}

growth_yr_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), y=inc.resid,
                                              #group=c(year, sp),
                                              #group=sp,
                                              #color=factor(sp),
                                              fill=factor(sp)))+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  geom_hline(yintercept=0)+
  facet_grid(.~year)+
  theme(axis.text.x = element_blank())+
  guides(color="none")

growth_yr_sp_detrend


```

```{r echo=F}

spag_all<-ggplot(dendro.ring, aes(x=year, y=inc.spline, group=treeID)) +
  geom_line()+ggtitle("fitted values (smooth spline)")

spag_all

spag_resid<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=treeID)) +
  geom_line()+ggtitle("residuals smooth spline")

spag_resid
```
## Detrending all species

```{r detrending all, echo=F}

#trying for one tree

dendro.data.noouts<-dendro.data.noouts %>%
  group_by(sp, treeID) %>%
  dplyr::mutate(n.obs=length(treeID))%>%
  filter(n.obs>=4)%>%
  ungroup()%>%
  filter(treeID!="HKK.92562")%>% #removing an issue with spline for now
  group_by(sp, treeID)%>%
  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))
```

For `r length(unique(dendro.data.noouts$sp))` species with dendrobands, all individual trees detrended for DBH with smooth splines

```{r echo=F}
growth_yr_detrend_all<-ggplot(dendro.data.noouts, aes(x=year, y=inc.resid, group=year))+ 
  #geom_violin()+ 
  geom_boxplot()+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))
growth_yr_detrend_all


```

## top 10 species

```{r echo=F}

dendro.by.sp<-dendro.data.noouts %>%
  group_by (sp, year) %>%
  dplyr::summarise(n.inds=length(sp),
    mean.inc = mean(inc.resid, na.rm=T),
                   median.inc = median(inc.resid, na.rm=T),
                   sd.inc = sd(inc.resid, na.rm=T),
    lower=mean.inc-sd.inc,
    upper=mean.inc+sd.inc)%>%
  filter(n.inds>=10)%>%
  arrange(year, -median.inc)

sp_top5<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA")

sp_top10<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA") 


dendro.top5<-dendro.by.sp %>%
  filter(sp %in% sp_top5)


dendro.top5.dat<-dendro.data.noouts %>%
  filter(sp %in% sp_top5)

dendro.top10.dat<-dendro.data.noouts %>%
  filter(sp %in% sp_top10)

hkk_census_top10_sp<-hkk_census %>%
  filter(CensusID==3)%>%
  filter(sp %in% sp_top10)

#pct agb of the top 15 dendro species
sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])

top10sp.map.census<-ggplot(hkk_census_top10_sp, aes(x=gx, y=gy, 
                                                    #size=as.numeric(dbh/10),
                                                    color=factor(sp, levels=sp_top10)))+
  geom_point(size=1)+
   scale_size(range = c(.1, 10), name="dbh")+
  guides(color=guide_legend(title="Species"))+
  scale_color_viridis_d()+
  ggtitle("top 10 species in 2004")+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

top10sp.map.census

trees.top10<-trees[trees$SPCODE.UPDATE %in% sp_top10,]


top10sp.map.dendro<-ggplot(trees.top10, aes(x=X, y=Y, 
                                                    #size=as.numeric(dbh/10), 
                                                    color=factor(SPCODE.UPDATE, levels=sp_top10)))+
  geom_point(size=1)+
   scale_size(range = c(.1, 10), name="dbh")+
  guides(color=guide_legend(title="Species"))+
  scale_color_viridis_d()+
  ggtitle("top 10 species in 2004")+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

top10sp.map.dendro

```

## Growth of most abundant species

```{r echo=F}

growth_sp5<-ggplot(dendro.top5.dat, aes(x=factor(sp, levels=sp_top5), y=diainc,
                                              fill=factor(sp, levels=sp_top5),
                                              color=factor(sp, levels=sp_top5)))+ 
  geom_violin(position = "dodge")+ 
    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top5.dat$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.top5.dat$diainc, na.rm=T), lty=2)+
  geom_text(aes(label=paste0("median=\n", round(median(dendro.top5.dat$diainc, na.rm=T),2), " cm"), 
                x=0.6, y=median(dendro.top5.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+
  xlab("")

growth_sp5

growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              fill=factor(sp, levels=sp_top10),
                                              color=factor(sp, levels=sp_top10)))+ 
  geom_violin(position = "dodge")+ 
    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_text(aes(label=paste0("median=\n", round(median(dendro.top10.dat$diainc, na.rm=T),2), " cm"), 
                x=0.6, y=median(dendro.top10.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+
  xlab("")

growth_sp10


growth_yr_sp<-ggplot(dendro.top5, aes(x=year, y=mean.inc, 
                                      color=sp))+ 
  geom_point()+
  geom_errorbar(aes(x=year, ymin=lower, ymax=upper))+
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))+
  facet_grid(sp~.)+
  guides(color="none")+
  geom_hline(yintercept = 0)+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  geom_vline(xintercept=c(2010, 2016), lty=2)
growth_yr_sp

#png("plots/growth_by_sp.png", height=8, width=4, units="in", res=300)
#growth_yr_sp
#dev.off()

```

## Size classes

```{r echo=F}

size_top10<-ggplot(dendro.top10.dat, aes(x=dbh))+
  geom_density()+
  geom_vline(aes(xintercept=35))+
  geom_vline(aes(xintercept=75))

size_top10

#trying three categories
# <50, 50-100, >100

dendro.top10.dat<-dendro.top10.dat %>%
  dplyr::mutate(size=ifelse(dbh<35, "< 35 cm dbh",
                       ifelse(dbh<70, "35-75 cm dbh", ">75 cm dbh")))


growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
      theme(axis.text.y = element_blank())+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+coord_flip()

growth_sp10

growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh")), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+

  geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge", col="black", fill="white", width=0.1)+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  #facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
      theme(axis.text.x = element_text(angle=45))+
  xlab("")+
  facet_grid(.~factor(sp, levels=sp_top10))+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))#+coord_flip()

growth_sp10

```

## Canopy illumination

```{r echo=F}

cii_top10<-ggplot(dendro.top10.dat, aes(x=cii))+
  geom_histogram()

cii_top10

#trying three categories
# <50, 50-100, >100

#dendro.top10.dat<-dendro.top10.dat %>%
#  dplyr::mutate(size=ifelse(dbh<50, "small",
#                       ifelse(dbh<100, "medium", "large")))


growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
                        aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(cii~.)+
      theme(axis.text.y = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+coord_flip()

growth_sp10_cii

growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
                        aes(
                          #x=factor(sp, levels=sp_top10), 
                          x=factor(cii), 
                            y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(.~factor(sp, levels = sp_top10))+
      theme(axis.text.y = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  #coord_flip()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))
growth_sp10_cii

```





## Top 15 species split by leaf habit


```{r evg/dec, echo=F}

sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA") 

#these 15 species had the most dendrobands and together made up 51.22% of the agb in 2009

#below values eyeballed from Williams et al. 2008 Fig 1.
sp_top15_phenoscore<-c(2.5, 0.8, 1.3, 4, 0.7, 
                       1, 1.25, 1.27, 0.2, 0.3,
                       0.4, 0.65, 3.95, 0.15, 0.25)

#no info on Alphonsea ventricosa in the paper. 
#Apparently evergreen, so used 1 as value - middle of evergreen range


pheno<-as.data.frame(cbind(sp_top15, sp_top15_phenoscore))
colnames(pheno)<-c("sp", "phenoscore")

dendro_top15<-dendro.data.noouts[dendro.data.noouts$sp %in% sp_top15,]
dendro_top15<-merge(dendro_top15, pheno, by="sp")
  

#categorising as evg and deciduous
dendro_top15$cat<-ifelse(dendro_top15$phenoscore<=2, "evg", "dec")

growth_yr_cat_detrend<-ggplot(dendro_top15, aes(x=factor(cat), y=inc.resid,
                                              #group=c(year, sp),
                                              #group=sp,
                                              fill=factor(cat)))+ 
  geom_violin(position = "dodge")+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  facet_grid(.~year)+
  geom_hline(yintercept = 0)+
  scale_fill_manual(values=c("sienna4", "darkgreen"))+
  theme(axis.text.x = element_blank())

growth_yr_cat_detrend

```

- identified and removed growth release
- hand pick high quality time series (10-20 per species maybe)
- try VH's model - temporal autocorr
- mean and median growth 
- by month plots - time difference
- different rates
